---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  out.height = "90%",
  fig.retina = 2
)
```

# specProc

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`specProc` package performs a number of pre-processing tasks commonly used in laser-induced breakdown spectroscopy (LIBS). Collectively, these are essential tools in LIBS calibration modeling, called chemometrics. These include:

-   peak analysis
-   Spectral-based normalization 
-   Smoothing and filtering
-   Robust or classical Box-Cox transformation 
-   Pearson or Spearman peaks correlation
-   Descriptive statistics such as mean, standard deviation and higher
    central moments 
-   Statistical tests
-   Rowwise and cellwise outliers detection
-   Robust or classical principal component analysis (PCA)
-   Orthogonal partial-least squares (OPLS)
-   PLS-based wavelength selection


## Installation

You can install the development version from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ChristianGoueguel/specProc")
```

## Peak analysis

The package peak analysis capabilities include:

-   Baseline removal
-   Peaks detection
-   Fitting single, multiple or overlapping peaks

### Baseline removal

For instance, when analyzing LIBS spectra, it is often more effective to subtract an estimated baseline from the data. The estimate is constructed by fitting a low-order polynomial function to the spectrum baseline. Then the resulting curve fit result is subtracted from the data.

```{r}
library(specProc)
```

```{r include=FALSE}
library(tidyverse)
library(patchwork)
```

```{r}
avg_spec <- arrow::read_parquet("~/Documents/Laserag/R&D projects/Soil/Eurofins/SOC/processed data/avg_spec")
LIBS_spec <- avg_spec %>%
  filter(spectra_id %in% "142146") %>%
  select(where(is.numeric))
```

```{r message=FALSE, warning=FALSE}
baseline_fit <- baseline(data = LIBS_spec)
str(baseline_fit, list.len = 5)  
```

```{r}
corrected_spec <- pluck(baseline_fit, "spec")
background <- pluck(baseline_fit, "bkg")
```

```{r include=FALSE}
background <- background %>%
  pivot_longer(
    cols = everything(), 
    names_to = "wavelength",
    values_to = "intensity"
    ) %>%
  modify_at("wavelength", as.numeric)
```

```{r}
plot1 <- plotSpec(data = LIBS_spec)
plot2 <- plotSpec(data = corrected_spec)
```



