---
title: "Adjusted Boxplot for Skewed Distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adjusted Boxplot for Skewed Distribution}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Skewness-adjusted boxplot is a variation of the traditional Turkey’s boxplot, which is a graphical representation of a dataset's distribution. The Turkey’s boxplot is particularly useful for visualizing the central tendency, spread, and potential outliers of a dataset. However, it does not convey information about the skewness, which is an important aspect of the data distribution. Skewness is a measure of the asymmetry of a dataset. When the distribution is positively skewed, the tail on the right side is longer, while a negatively skewed distribution has a longer tail on the left side. A skewness-adjusted boxplot modifies the traditional boxplot to provide a visual representation of the skewness in addition to the other summary statistics.

The skewness-adjusted boxplot using the medcouple (MC) statistic provides a more accurate and robust representation of the data distribution. It was introduced by Brys et al. (J. Comput. Graph. Stat. 13, 2004) and Hubert et al. (Comput. Stat. Data Anal., 52, 2008). The medcouple is based on the concept of "coupling," where it measures the degree of asymmetry around the median of a dataset. The MC statistic ranges from -1 to 1, where a positive value indicates a right-skewed distribution and a negative value indicates a left-skewed distribution. A value close to zero suggests a symmetric distribution.

```{r setup}
ssh = suppressPackageStartupMessages
ssh(library(tidyverse))
```

```{r}
source("~/Documents/Packages/specProc/R/adj_boxplot.R")
```

```{r}
data(glass, package = "chemometrics")
data(glass.grp, package = "chemometrics")
```

```{r}
df_tbl <-
  cbind(glass.grp, glass) %>%
  as_tibble() %>%
  rename(group = glass.grp) %>%
  modify_at("group", as_factor)
```

```{r}
df_tbl %>% slice_head(n = 5)
```

```{r message=FALSE, warning=FALSE}
adjBox <- df_tbl %>% select(where(is.numeric)) %>% adj_boxplot()
```

```{r}
adjBox %>% pluck("adj_box_stat")
```

```{r fig.height=7, fig.width=9}
adjBox %>% pluck("plot")
```

