---
title: "Predicting soil clay content from LIBS spectra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicting soil clay content from LIBS spectra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  dpi = 96
)
```

Preprocessing is only useful if it improves the analysis it serves. This
vignette uses a concrete goal to evaluate preprocessing choices:
predicting the clay content of the 50 soil samples in `specLIBS` with
partial least squares (PLS) regression.

The main difficulty is statistical, not spectroscopic. With 50 samples
and 7152 variables, a model can fit the calibration data almost
perfectly, so everything depends on estimating prediction error honestly.
Several common shortcuts bias that estimate downward. The vignette
quantifies two of them.

```{r setup}
library(specProc)
data(specLIBS)

meta <- specLIBS[1:8]
X <- as.matrix(specLIBS[-(1:8)])
```

## The target

```{r target, fig.height = 3.5}
clay <- tapply(meta$Clay, meta$Sample, unique)
summary(clay)
hist(clay, breaks = 15, col = "grey80", border = "white", main = NULL,
     xlab = "Clay content (%)")
rug(clay)
```

The distribution is left-skewed. Most samples contain 20 to 45% clay,
while a handful of samples lie well above 55%. These few samples have high
leverage: the model's behaviour at high clay content rests on very little
data, and any error summary is sensitive to how they fall in the
cross-validation folds.

## Preprocessing without leakage

Preprocessing steps fall into two groups:

-   **Steps computed from each spectrum alone** (`baseline_arpls()`,
    `snv()`, `normalize()`): they use no information from other samples, so
    they can be applied once to the whole data set.
-   **Steps estimated from a set of samples** (the MSC reference spectrum,
    EPO and GLSW filters, the OSC family, centering and scaling): they must
    be estimated on the calibration part of each cross-validation split
    only, and then applied to the held-out part.

We apply the per-spectrum steps now, and average the 8 shots of each sample:

```{r preprocess}
Xb <- as.matrix(baseline_arpls(X, lambda = 1e5, max.iter = 20)$correction)
Xsnv <- as.matrix(snv(Xb)$correction)

sample_mean <- function(M) {
  a <- average(cbind(Sample = meta$Sample, as.data.frame(M)), Sample)
  out <- as.matrix(a[-1])
  rownames(out) <- a$Sample
  out
}
raw_s <- sample_mean(X)
base_s <- sample_mean(Xb)
snv_s <- sample_mean(Xsnv)
y <- clay[rownames(snv_s)]
```

## Validation design

We use **repeated 5-fold cross-validation** with the sample as the unit.
Each sample is one row after averaging, so all 8 shots of a sample are
always on the same side of a split. Within each calibration set, the
number of PLS components is chosen by an inner cross-validation (up to
10 components). The outer folds therefore never influence this choice:
this is nested cross-validation. The whole procedure is repeated 5 times
with different fold assignments, to show how much the error estimate
itself varies.

```{r helpers}
fit_predict <- function(Xtr, ytr, Xte, ncomp = NULL, max_comp = 10) {
  if (is.null(ncomp)) {
    inner <- pls::plsr(ytr ~ Xtr, ncomp = max_comp, validation = "CV", segments = 5)
    ncomp <- which.min(inner$validation$PRESS[1, ])
  }
  model <- pls::plsr(ytr ~ Xtr, ncomp = ncomp)
  list(pred = drop(predict(model, newdata = Xte, ncomp = ncomp)), ncomp = ncomp)
}

# Repeated K-fold CV. `pipeline(train, test)` returns the preprocessed
# calibration and validation matrices and, optionally, a fixed ncomp.
cross_validate <- function(pipeline, repeats = 5, K = 5, seed = 2024) {
  set.seed(seed)
  n <- length(y)
  preds <- matrix(NA_real_, n, repeats)
  ncomps <- integer(0)
  for (r in seq_len(repeats)) {
    folds <- sample(rep(seq_len(K), length.out = n))
    for (k in seq_len(K)) {
      test <- which(folds == k)
      train <- which(folds != k)
      d <- pipeline(train, test)
      f <- fit_predict(d$train, y[train], d$test, ncomp = d$ncomp)
      preds[test, r] <- f$pred
      ncomps <- c(ncomps, f$ncomp)
    }
  }
  rmse <- apply(preds, 2, function(p) sqrt(mean((p - y)^2)))
  r2 <- apply(preds, 2, function(p) 1 - sum((p - y)^2) / sum((y - mean(y))^2))
  list(preds = preds, rmse = rmse, r2 = r2, ncomp = stats::median(ncomps))
}
```

## Comparing preprocessing pipelines

We compare six pipelines. The same seed gives every pipeline the same
fold assignments, so the comparisons are paired:

```{r pipelines}
pipelines <- list(
  `raw counts` = function(tr, te) list(train = raw_s[tr, ], test = raw_s[te, ]),

  `baseline` = function(tr, te) list(train = base_s[tr, ], test = base_s[te, ]),

  `baseline + SNV` = function(tr, te) list(train = snv_s[tr, ], test = snv_s[te, ]),

  # MSC: the reference spectrum is estimated on the calibration samples only
  `baseline + MSC` = function(tr, te) {
    cal <- msc(base_s[tr, ])
    val <- msc(base_s[te, ], xref = cal$reference)
    list(train = as.matrix(cal$correction), test = as.matrix(val$correction))
  },

  # EPO: clutter = shot-to-shot deviations of the calibration samples only
  `SNV + EPO` = function(tr, te) {
    shots <- meta$Sample %in% rownames(snv_s)[tr]
    S <- Xsnv[shots, ]
    clutter <- S - apply(S, 2, function(v) ave(v, meta$Sample[shots]))
    filtered <- as.matrix(epo(snv_s, ncomp = 3, clutter = clutter)$correction)
    list(train = filtered[tr, ], test = filtered[te, ])
  },

  # OPLS filter (projected OSC) fitted on the calibration samples only,
  # followed by a one-component PLS model
  `SNV + OPLS filter` = function(tr, te) {
    f <- projected_osc(snv_s[tr, ], y[tr], ncomp = 4, newdata = snv_s[te, ])
    list(train = as.matrix(f$correction), test = as.matrix(f$newdata$correction), ncomp = 1)
  }
)

results <- lapply(pipelines, cross_validate)
```

```{r results}
null_rmse <- sqrt(mean((y - mean(y))^2))
summary_table <- data.frame(
  pipeline = names(results),
  RMSE = sapply(results, function(r) mean(r$rmse)),
  RMSE_sd = sapply(results, function(r) sd(r$rmse)),
  R2 = sapply(results, function(r) mean(r$r2)),
  ncomp = sapply(results, function(r) r$ncomp),  # median number of PLS components
  row.names = NULL
)
summary_table[2:4] <- round(summary_table[2:4], 2)
summary_table
c(null_model_RMSE = round(null_rmse, 2))
```

`RMSE` and `R2` are averages over the 5 repeats, and `RMSE_sd` is the
standard deviation of the RMSE across repeats. All pipelines predict clay
content far better than the null model, which
predicts the mean for every sample (RMSE `r round(null_rmse, 1)`%). With
an RMSE of about 8% and a cross-validated R² of about 0.7, the spectra
explain most of the variation in clay content.

Are the pipelines different from each other? The fold assignments are
shared, so we compare each pipeline with the baseline-only pipeline
repeat by repeat:

```{r paired}
reference <- results$baseline$rmse
paired <- t(sapply(results, function(r) {
  d <- r$rmse - reference
  c(mean_difference = mean(d), min = min(d), max = max(d))
}))
round(paired, 2)
```

Every pipeline differs from the baseline-only pipeline by less than 0.25
points of RMSE on average, and for every pipeline the sign of the
difference changes from one repeat to another. With 50 samples, these data cannot distinguish the pipelines
reliably. This matches the finding of the preprocessing vignette: SNV
improves the repeatability of the spectra, but removes between-sample
variation too, and the two effects roughly cancel for clay prediction.

The honest conclusion is that preprocessing choice matters little here.
A claim that one pipeline is better would need more samples, or an
independent validation set.

## How much optimism do common shortcuts add?

### Shortcut 1: splitting replicates across folds

If the 400 shots are cross-validated as if they were independent, shots
of the same sample end up in both the calibration and the validation
folds. The model is then partly validated on the samples it was trained
on:

```{r leaky-shots}
set.seed(2024)
y_shot <- meta$Clay
shot_folds <- sample(rep(1:5, length.out = nrow(Xsnv)))
pred_shot <- numeric(nrow(Xsnv))
for (k in 1:5) {
  test <- shot_folds == k
  pred_shot[test] <- fit_predict(Xsnv[!test, ], y_shot[!test], Xsnv[test, ])$pred
}
shot_rmse <- sqrt(mean((pred_shot - y_shot)^2))
c(shot_level_CV = round(shot_rmse, 2),
  sample_level_CV = round(mean(results$`baseline + SNV`$rmse), 2))
```

Shot-level cross-validation reports an RMSE about
`r round(100 * (1 - shot_rmse / mean(results[["baseline + SNV"]]$rmse)))`%
lower than the sample-level estimate. The optimism is moderate here,
because the shots of a sample differ appreciably. It grows as replicates
become more alike, and it can be large for homogeneous materials.

### Shortcut 2: fitting a supervised filter before cross-validating

The OPLS filter uses the response `y` to decide what to remove. If it is
fitted once on all 50 samples and cross-validation is run afterwards, the
validation samples have already influenced the filter:

```{r leaky-filter}
filtered_all <- as.matrix(projected_osc(snv_s, y, ncomp = 4)$correction)
leaky <- cross_validate(function(tr, te) {
  list(train = filtered_all[tr, ], test = filtered_all[te, ], ncomp = 1)
})
c(filter_before_CV = round(mean(leaky$rmse), 2),
  filter_inside_CV = round(mean(results$`SNV + OPLS filter`$rmse), 2))
```

Fitting the filter before cross-validation gives an RMSE
`r round(100 * (1 - mean(leaky$rmse) / mean(results[["SNV + OPLS filter"]]$rmse)))`%
lower than the honest estimate, a far larger effect than any difference
between the pipelines above. The same bias affects `osc()`,
`direct_osc()`, `direct_orthogonal()`, `nas()` and `o2pls()`, and any
variable selection that uses the response. Refit these steps inside
every fold, including the inner loop that chooses the number of
components.

## Looking at the predictions

The out-of-fold predictions, averaged over the 5 repeats, show where the
model succeeds and where it fails:

```{r predictions, fig.height = 4.5, fig.width = 5, out.width = "70%"}
pred <- rowMeans(results$`baseline + SNV`$preds)
plot(y, pred, pch = 21, bg = "grey70", xlim = c(0, 85), ylim = c(0, 85),
     xlab = "Measured clay (%)", ylab = "Cross-validated prediction (%)")
abline(0, 1, lty = 2)
```

Three of the four samples above 55% clay are underpredicted, by 12 to 21
points. With so few calibration samples in that range, the model pulls
extreme predictions toward the mean, and it should not be used to predict
clay contents above about 55% without more calibration data there. At the
other end, one sample receives a negative prediction: a linear model does
not respect the bounds of a percentage.

## Further considerations

-   **Clay, sand and silt are compositional.** They sum to 100% for every
    sample, so separate models for each fraction will not, in general,
    predict fractions that sum to 100%. To predict all three, model
    log-ratios such as \(\log(\text{clay}/\text{silt})\) and
    \(\log(\text{sand}/\text{silt})\), and transform the predictions back.
    This also keeps predictions within 0–100%.
-   **Cross-validation estimates error for similar samples.** It says
    nothing about soils from other regions, other instruments or other
    measurement sessions. An independent test set measured later is
    the stronger evidence, and `pds()` or `glsw()` can help when a model
    must be transferred between instruments.
-   **Report the variability of the error estimate.** A single RMSE from a
    single cross-validation split hides the spread shown in the table above.
