---
title: "Removing unwanted variation: a comparison of orthogonalization methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Removing unwanted variation: a comparison of orthogonalization methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  dpi = 96
)
```

Most of the variation in a set of LIBS spectra has nothing to do with the
property we want to predict. Plasma conditions change the total emitted
intensity from shot to shot and sample to sample, and every element in the
sample adds its own lines. Orthogonalization methods try to identify this
unwanted variation and remove it before calibration.

specProc implements two families of such methods, which differ in how
they define "unwanted":

| Family | Unwanted variation is… | Functions |
|:-------|:-----------------------|:----------|
| External | the variation between spectra that should be identical (clutter) | `epo()`, `glsw()` |
| Response-orthogonal | the variation in $\mathbf{X}$ uncorrelated with the response $\mathbf{y}$ | `osc()`, `direct_orthogonal()`, `nas()`, `direct_osc()`, `projected_osc()`, `o2pls()`, `opls()`, `y_gradient_glsw()` |

This vignette applies all of them to the same problem, predicting the
potassium content of forage samples, and asks three questions:

1.  **What does each method remove?**
2.  **Does removing it improve predictions** on samples not used to fit
    the filter?
3.  **Which methods are actually the same method?**

```{r setup}
library(specProc)
data(fourrage)

meta <- fourrage[1:14]
X <- as.matrix(fourrage[-(1:14)])
wl <- as.numeric(colnames(X))
dim(X)
```

`fourrage` contains 368 measurements of 365 forage samples (three samples
were measured twice). Each measurement is the mean of 8 laser shots. The
target is potassium, which is available for every measurement and spans
`r round(min(meta$K), 1)` to `r round(max(meta$K), 1)`%.

## Preparing the spectra

Two practical steps come first:

-   **Channel overlap.** Two spectrometers overlap near 766 nm, so a few
    channels repeat wavelengths already covered. We drop them.
-   **Binning.** The GLSW filters are $p \times p$ matrices: with 7152
    channels, each filter would take about 400 MB. Summing 4 adjacent
    channels reduces this to 25 MB and speeds up every method, at the cost
    of some spectral resolution. Most emission lines in these spectra span
    several channels, so little information is lost.

Finally, SNV normalizes each spectrum.

```{r bin}
keep <- wl > cummax(c(-Inf, wl[-length(wl)]))
sum(!keep)  # overlapping channels dropped
bin <- ceiling(seq_len(sum(keep)) / 4)
Xb <- t(rowsum(t(X[, keep]), bin))
wlb <- as.vector(tapply(wl[keep], bin, mean))
Xs <- as.matrix(snv(Xb)$correction)
dim(Xs)
```

## Identifying the unwanted variation

A principal component analysis shows where the variance lies. We correlate
the scores of the first five components with the total emitted intensity
and with the reference contents of several elements:

```{r pca}
y <- meta$K
total <- rowSums(X)
chem <- as.data.frame(meta[c("K", "Ca", "Mg", "Na", "P", "S")])
pc_raw <- prcomp(Xb)
pc_snv <- prcomp(Xs)
explained <- function(pc, k = 5) round(100 * pc$sdev[1:k]^2 / sum(pc$sdev^2), 1)
rbind(raw = explained(pc_raw), snv = explained(pc_snv))  # % variance, PC1-PC5

score_correlations <- function(scores) {
  round(cor(scores, cbind(total, chem), use = "pairwise.complete.obs"), 2)
}
score_correlations(pc_raw$x[, 1:5])
score_correlations(pc_snv$x[, 1:5])
```

The raw spectra are dominated by one structure. The first component
explains `r explained(pc_raw)[1]`% of the variance, and its scores are
almost perfectly correlated with the total intensity. This is the
multiplicative effect of the plasma: a hotter or denser plasma raises all
lines together. SNV removes most of it, but the SNV spectra still carry
several structures that have little to do with potassium:

-   **PC1 and PC2** still follow the total intensity, and also sodium and
    magnesium. After SNV, intensity changes appear as changes in line
    ratios, which a per-spectrum scaling cannot remove.
-   **PC3** contrasts potassium with calcium, the main cation that varies
    in the opposite direction.

How much of the SNV variance is related to potassium at all? Regressing
each channel on $\mathbf{y}$ gives the share of variance that a linear
function of potassium content explains:

```{r y-share}
Xc <- scale(Xs, scale = FALSE)
yc <- y - mean(y)
y_part <- yc %*% crossprod(yc, Xc) / sum(yc^2)
round(100 * sum(y_part^2) / sum(Xc^2), 1)  # % of the variance
```

Only about `r round(100 * sum(y_part^2) / sum(Xc^2))`% of the variance is
directly related to potassium. The remaining variance is the
target of the orthogonalization methods.

## Calibration and test sets

We set aside one third of the samples as a test set. Filters and PLS models
are always fitted on the calibration set only, and every tuning decision
uses cross-validation within the calibration set.

The external methods need clutter: spectra that should be identical but
are not. The three samples measured twice provide this. Their difference
spectra describe how the measurement of a given material varies from one
session to the next. We keep these three samples in the calibration set.

```{r split}
set.seed(1)
samples <- unique(meta$Sample)
twice <- unique(meta$Sample[duplicated(meta$Sample)])
test_samples <- sample(setdiff(samples, twice), round(length(samples) / 3))
test <- meta$Sample %in% test_samples
cal <- !test
c(calibration = sum(cal), test = sum(test))

first <- match(twice, meta$Sample)
second <- sapply(twice, function(s) max(which(meta$Sample == s)))
X1 <- Xs[first, ]
X2 <- Xs[second, ]
clutter <- X2 - X1

Xcal <- Xs[cal, ]
ycal <- y[cal]
```

Three difference spectra are a small clutter matrix. In practice, repeated
measurements of a few reference materials over several sessions would
describe this variation better.

## A common interface for the filters

Every method is estimated on calibration spectra and then applied, with
the same parameters, to new spectra. `epo()`, `osc()`,
`direct_orthogonal()`, `direct_osc()`, `projected_osc()` and `o2pls()`
return a fitted filter, and `predict(filter, newdata)` corrects new spectra
with it (see `?predict.specproc_filter`). `glsw()` and `y_gradient_glsw()`
return the filter matrix $\mathbf{G}$, which is applied as
$\mathbf{X}_{new}\mathbf{G}$.

To compare the methods, we wrap each one in a function `fit(X, y, k)` that
returns a function to correct new spectra. `k` is the tuning parameter:
the number of components removed or, for GLSW, the strength of the
down-weighting.

```{r filters}
# A function that corrects new spectra with a fitted filter
corrector <- function(filter) function(Xn) as.matrix(predict(filter, Xn))
mat <- function(x) as.matrix(x)

filters <- list(
  `none` = list(grid = 0, fit = function(X, y, k) identity),

  # External: the clutter directions are projected out ...
  `EPO` = list(grid = 1:3, fit = function(X, y, k) {
    corrector(epo(X, ncomp = k, clutter = clutter))
  }),
  # ... or down-weighted; alpha is set relative to the largest clutter eigenvalue
  `GLSW` = list(grid = 0:4, fit = function(X, y, k) {
    lambda <- svd(scale(clutter, scale = FALSE), 0, 0)$d[1]^2
    G <- mat(glsw(X1, X2, alpha = lambda * 10^-k))
    function(Xn) Xn %*% G
  }),

  # Response-orthogonal
  `OSC (Wold)` = list(grid = 1:4, fit = function(X, y, k) {
    corrector(osc(X, y, method = "wold", ncomp = k))
  }),
  `OSC (Sjoblom)` = list(grid = 1:4, fit = function(X, y, k) {
    corrector(osc(X, y, method = "sjoblom", ncomp = k))
  }),
  `OSC (Fearn)` = list(grid = 1:4, fit = function(X, y, k) {
    corrector(osc(X, y, method = "fearn", ncomp = k))
  }),
  `DO / NAS` = list(grid = 1:4, fit = function(X, y, k) {
    corrector(direct_orthogonal(X, y, ncomp = k))
  }),
  `DOSC` = list(grid = 1:4, fit = function(X, y, k) {
    corrector(direct_osc(X, y, ncomp = k))
  }),
  `POSC / OPLS` = list(grid = 1:4, fit = function(X, y, k) {
    corrector(projected_osc(X, y, ncomp = k + 1))  # k orthogonal components
  }),
  # alpha relative to the largest eigenvalue of differences between neighbours in y
  `y-gradient GLSW` = list(grid = 0:4, fit = function(X, y, k) {
    lambda <- svd(diff(X[order(y), ]), 0, 0)$d[1]^2
    G <- mat(y_gradient_glsw(X, y, alpha = lambda * 10^-k))
    function(Xn) Xn %*% G
  })
)
```

`opls()` is not included: it wraps the Bioconductor package ropls and
returns a fitted model rather than a filter. `projected_osc()` gives the
same filtered data, as shown below.

## What does each method remove?

We fit each method to the calibration set with a comparable setting (2
components, or $\alpha$ at 1% of the largest eigenvalue for GLSW). For the
removed part of the spectra, $\mathbf{X} - f(\mathbf{X})$, we compute:

-   **`removed`:** its share of the calibration variance, in percent.
-   **The correlations of its first principal component** with the total
    intensity and with the element contents (absolute values).

```{r what-removed}
k_show <- c(EPO = 2, GLSW = 2, `OSC (Wold)` = 2, `OSC (Sjoblom)` = 2,
            `OSC (Fearn)` = 2, `DO / NAS` = 2, DOSC = 2, `POSC / OPLS` = 2,
            `y-gradient GLSW` = 2)
direction <- list()
removed <- t(sapply(names(k_show), function(m) {
  f <- filters[[m]]$fit(Xcal, ycal, k_show[[m]])
  R <- scale(Xcal - f(Xcal), scale = FALSE)
  s <- svd(R, nu = 1, nv = 1)
  direction[[m]] <<- s$v[, 1]
  c(removed = 100 * sum(R^2) / sum(scale(Xcal, scale = FALSE)^2),
    abs(cor(s$u, cbind(total = total[cal], chem[cal, ]), use = "pairwise.complete.obs"))[1, ])
}))
round(removed, 2)
```

The table shows what each family targets:

-   **The response-orthogonal methods remove the most variance**, about
    60–70% with only two components. The removed part is uncorrelated with
    potassium, exactly so for Wold's and Fearn's OSC, DOSC and POSC. DO/NAS
    and Sjöblom's OSC keep a small correlation, because they remove the
    principal directions of the $\mathbf{y}$-orthogonal space, and the
    scores of $\mathbf{X}$ on these directions need not be orthogonal to
    $\mathbf{y}$. What these methods remove is strongly related to the
    total intensity and to sodium, the structures found by the PCA above.
-   **EPO and GLSW remove less**, and what they remove is mostly the
    total intensity (correlations of 0.84 and 0.90). When a sample is measured
    again, what changes most is the plasma, not the composition. It also
    shows that SNV does not remove all intensity effects.
-   **y-gradient GLSW is the gentlest filter:** it removes about 10% of the
    variance, again related to sodium and intensity.

The first direction of the removed part (the loading) shows which
emission lines are involved:

```{r directions, fig.height = 7}
lines_nm <- c(`Mg` = 279.55, `Ca` = 393.37, `Na` = 588.99, `H` = 656.28,
              `K` = 766.49, `O` = 777.19)
op <- par(mfrow = c(4, 1), mar = c(2, 4, 1.5, 1), oma = c(2, 0, 0, 0))
for (m in c("EPO", "DO / NAS", "POSC / OPLS", "y-gradient GLSW")) {
  plot(wlb, direction[[m]], type = "l", xlab = "", ylab = "Loading", main = m,
       cex.main = 0.9)
  abline(v = lines_nm, col = "grey60", lty = 3)
  mtext(names(lines_nm), side = 3, at = lines_nm, cex = 0.6, line = 0)
}
mtext("Wavelength (nm)", side = 1, outer = TRUE, line = 0.5)
par(op)
```

The sign of each loading is arbitrary. The four directions share a common
pattern: the emission lines of the sample's metals (Mg, Ca and above all
the Na doublet at 589 nm) vary against the O I line at 777 nm and the
continuum above 800 nm. For a measurement in air, the O I line comes from
both the atmosphere and the organic matrix, so this contrast likely
reflects how much material is ablated and how the plasma develops rather
than the composition of the sample. The sodium doublet is the largest feature for the
response-orthogonal methods and y-gradient GLSW: its variation between
samples is large and independent of potassium. The potassium lines
barely appear in these directions.

## Does removing it improve predictions?

We now combine each filter with PLS regression and tune both the filter
parameter and the number of PLS components (up to 15) by 5-fold
cross-validation within the calibration set. The two measurements of a
sample are always in the same fold. The filter is refitted in every fold,
so the validation spectra never influence it.

```{r cv-helpers}
max_comp <- 15
rmse <- function(pred, obs) sqrt(mean((pred - obs)^2))
pls_path <- function(Xtr, ytr, Xte) {
  model <- pls::plsr(ytr ~ Xtr, ncomp = max_comp, method = "simpls")
  matrix(predict(model, newdata = Xte, ncomp = seq_len(max_comp)), nrow(Xte))
}

set.seed(2)
cal_samples <- unique(meta$Sample[cal])
fold_of <- setNames(sample(rep(1:5, length.out = length(cal_samples))), cal_samples)
folds <- fold_of[meta$Sample[cal]]

tune_filter <- function(filter) {
  cv <- sapply(filter$grid, function(k) {
    pred <- matrix(NA_real_, sum(cal), max_comp)
    for (f in 1:5) {
      v <- folds == f
      clean <- filter$fit(Xcal[!v, ], ycal[!v], k)
      pred[v, ] <- pls_path(clean(Xcal[!v, ]), ycal[!v], clean(Xcal[v, , drop = FALSE]))
    }
    apply(pred, 2, rmse, ycal)
  })
  cv <- matrix(cv, max_comp)
  best <- which(cv == min(cv), arr.ind = TRUE)[1, ]
  list(k = filter$grid[best[2]], ncomp = unname(best[1]), rmsecv = min(cv))
}
```

After tuning, each filter and its PLS model are refitted to the whole
calibration set and applied to the test set:

```{r compare}
tuned <- lapply(filters, tune_filter)
test_pred <- sapply(names(filters), function(m) {
  clean <- filters[[m]]$fit(Xcal, ycal, tuned[[m]]$k)
  pls_path(clean(Xcal), ycal, clean(Xs[test, ]))[, tuned[[m]]$ncomp]
})
```

With 122 test spectra, the test error itself is uncertain. Because all
methods predict the same test samples, we compare each one with plain PLS
(`none`) using a paired bootstrap of the test samples. `diff_low` and
`diff_high` bound a 95% interval for the difference in RMSEP; negative
values favor the filter.

```{r compare-table}
set.seed(3)
boot <- replicate(2000, {
  i <- sample(nrow(test_pred), replace = TRUE)
  apply(test_pred[i, ], 2, rmse, y[test][i]) - rmse(test_pred[i, "none"], y[test][i])
})
comparison <- data.frame(
  filter_k = sapply(tuned, `[[`, "k"),
  pls_ncomp = sapply(tuned, `[[`, "ncomp"),
  RMSECV = sapply(tuned, `[[`, "rmsecv"),
  RMSEP = apply(test_pred, 2, rmse, y[test]),
  diff_low = apply(boot, 1, quantile, 0.025),
  diff_high = apply(boot, 1, quantile, 0.975)
)
round(comparison, 3)
c(null_RMSEP = round(rmse(mean(ycal), y[test]), 3))
```

All models predict potassium far better than the null model, which
predicts the calibration mean for every sample. The comparison with plain
PLS is clear:

-   **No filter improves predictions measurably.** Every interval for the
    difference with plain PLS includes zero or favors plain PLS. The
    differences in RMSEP are a few hundredths of a percent of potassium,
    against a typical error of about
    `r round(comparison["none", "RMSEP"], 2)`%.
-   **Filtering mostly moves components from PLS to the filter.** DOSC with
    one component needs fewer PLS components, and Wold's OSC and POSC
    remove three components and need three fewer. The total model
    complexity hardly changes.
-   **Several filtered models are plain PLS in disguise.** The OSC (Fearn)
    and POSC models give exactly the same test predictions as plain PLS
    (both bootstrap bounds are 0), as the next section explains.
-   **y-gradient GLSW gives the most parsimonious model**, with
    `r comparison["y-gradient GLSW", "pls_ncomp"]` PLS components, and
    the lowest RMSEP, but the improvement is not distinguishable from
    chance. It is the one method here that reshapes the covariance of
    $\mathbf{X}$ rather than removing whole directions, which can make the
    remaining PLS model simpler.

The external filters could not help much: their clutter comes from three
samples, and the main thing it describes, the intensity effect, is
variation PLS already learns to ignore from 246 calibration spectra.

## Which methods are the same method?

Several functions compute the same filter, or a filter that cannot change
PLS predictions:

```{r equivalences}
# NAS and DO: both project X onto the complement of the leading principal
# directions of X after removing the part explained by y
all.equal(mat(nas(Xcal, ycal, ncomp = 2)),
          mat(direct_orthogonal(Xcal, ycal, ncomp = 2)$correction),
          check.attributes = FALSE)

# POSC and O2PLS with a single response: both give OPLS-filtered data
all.equal(mat(projected_osc(Xcal, ycal, ncomp = 3)$correction),
          mat(o2pls(Xcal, ycal, ncomp = 1, nx = 2)$correction),
          check.attributes = FALSE)

# OPLS filter with 2 orthogonal components + 1-component PLS
# gives the same predictions as a 3-component PLS model
clean <- filters$`POSC / OPLS`$fit(Xcal, ycal, 2)
all.equal(pls_path(clean(Xcal), ycal, clean(Xs[test, ]))[, 1],
          pls_path(Xcal, ycal, Xs[test, ])[, 3])
```

The first two results mean that `nas()` and `direct_orthogonal()`, and
`projected_osc()` and `o2pls()` (with one response), are
interchangeable. The third explains the zero differences in the
comparison table. OPLS splits the systematic variation of a PLS model into
a predictive and an orthogonal part, but the model is the same (Trygg and
Wold, 2002; Kemsley and Tapp, 2009). The same holds for Fearn's OSC in this
comparison. These filters are still useful for interpretation: the
orthogonal components show what varies in the spectra independently of
potassium, as in the figure above.

## The main pitfall: fitting the filter before validating

The response-orthogonal filters use $\mathbf{y}$. If a filter is fitted
once on all calibration samples and cross-validation is run afterwards,
the validation samples have already shaped the filter. DOSC shows the
consequence in its most extreme form:

```{r leak}
leak <- t(sapply(c(1, 2, 4, 8), function(k) {
  clean <- corrector(direct_osc(Xcal, ycal, ncomp = k))
  Xf <- clean(Xcal)
  pred <- matrix(NA_real_, sum(cal), max_comp)
  for (f in 1:5) {
    v <- folds == f
    pred[v, ] <- pls_path(Xf[!v, ], ycal[!v], Xf[v, , drop = FALSE])
  }
  cv <- apply(pred, 2, rmse, ycal)
  a <- which.min(cv)
  c(filter_k = k, pls_ncomp = a, leaky_RMSECV = min(cv),
    RMSEP = rmse(pls_path(Xf, ycal, clean(Xs[test, ]))[, a], y[test]))
}))
round(leak, 3)
```

With the filter fitted before cross-validation, the cross-validated error
is close to zero, while the error on the test set is higher than for plain
PLS. DOSC first projects $\mathbf{y}$ onto the space of $\mathbf{X}$. With
far more channels than samples, this projection reproduces the calibration
values almost exactly, and so does the filtered $\mathbf{X}$. The
calibration vignette (`vignette("calibration", package = "specProc")`)
shows a milder version of the same bias for the OPLS filter. The rule is
the same for every method in the response-orthogonal family: refit the
filter inside every cross-validation fold.

## The same analysis with tidymodels

Each filter is also available as a recipe step for the tidymodels
framework: `step_epo()`, `step_glsw()`, `step_osc()`,
`step_direct_orthogonal()`, `step_direct_osc()`, `step_projected_osc()` and
`step_y_gradient_glsw()`. In a workflow, the steps are re-estimated on the
analysis set of every resample, so supervised filters cannot leak
information from the assessment set, and the filter parameters can be tuned
together with the model. The preprocessing steps `step_baseline()`,
`step_snv()`, `step_msc()`, `step_emsc()`, `step_pareto_scale()` and
`step_poisson_scale()` can precede them in the same recipe.

The code below needs recipes, parsnip, workflows, tune, rsample,
yardstick, dials, plsmod and mixOmics (from Bioconductor), and is only run
when they are installed. It tunes Fearn's OSC filter and the PLS model
together, with the samples as the resampling unit. The two `num_comp`
parameters get distinct ids.

```{r tidymodels, message = FALSE, eval = all(vapply(c("recipes", "parsnip", "workflows", "tune", "rsample", "yardstick", "dials", "plsmod", "mixOmics"), requireNamespace, logical(1), quietly = TRUE))}
library(recipes)
library(parsnip)
library(workflows)
library(tune)

spectra <- data.frame(K = y, Sample = meta$Sample, Xs)
cal_data <- spectra[cal, ]
test_data <- spectra[test, ]

rec <- recipe(K ~ ., data = cal_data) |>
  update_role(Sample, new_role = "id") |>
  step_osc(all_predictors(), method = "fearn", num_comp = tune("filter_comp"))
# mixOmics scales every channel to unit variance by default; scale = FALSE
# only centers, like pls::plsr() above
model <- pls(num_comp = tune()) |>
  set_mode("regression") |>
  set_engine("mixOmics", scale = FALSE)
wf <- workflow(rec, model)

set.seed(4)
resamples <- rsample::group_vfold_cv(cal_data, group = Sample, v = 5)
grid <- expand.grid(filter_comp = 1:3, num_comp = 1:12)
tuned_wf <- tune_grid(wf, resamples = resamples, grid = grid,
                      metrics = yardstick::metric_set(yardstick::rmse))
show_best(tuned_wf, metric = "rmse", n = 3)

final <- finalize_workflow(wf, select_best(tuned_wf, metric = "rmse")) |>
  fit(data = cal_data)
c(RMSEP = rmse(predict(final, test_data)$.pred, test_data$K))
```

The best settings trade filter components against PLS components, with
the same total and the same cross-validated error, and the test error of
the final workflow matches the plain PLS and OSC (Fearn) rows of the
comparison table. The fold assignments differ from those used above, so
the chosen settings can differ as well.

## Summary

-   **Identify first.** A PCA with correlations to known factors (total
    intensity, other elements) shows what the unwanted variation is. Here,
    it is mainly the plasma intensity effect and the variation of sodium
    and calcium.
-   **External methods remove what you tell them to.** EPO and GLSW are
    only as good as the clutter. Differences between repeated measurements
    of the same material captured the intensity effect, which SNV leaves
    in part.
-   **Response-orthogonal methods remove a lot of variance, but add no
    information.** They make models smaller and easier to interpret, not
    more accurate: the removed variation is variation that PLS already
    ignores. Some (OPLS/POSC, O2PLS, Fearn's OSC here) give exactly the
    PLS predictions.
-   **Validate honestly.** Supervised filters must be refitted inside each
    fold. Otherwise the cross-validated error can be meaningless.

## References

-   Andersson, C.A. (1999). Direct orthogonalization. *Chemometrics and
    Intelligent Laboratory Systems*, 47(1):51–63.
-   Fearn, T. (2000). On orthogonal signal correction. *Chemometrics and
    Intelligent Laboratory Systems*, 50(1):47–52.
-   Kemsley, E.K., Tapp, H.S. (2009). OPLS filtered data can be obtained
    directly from non-orthogonalized PLS1. *Journal of Chemometrics*,
    23(5):263–264.
-   Martens, H., Høy, M., Wise, B.M., Bro, R., Brockhoff, P.B. (2003).
    Pre-whitening of data by covariance-weighted preprocessing. *Journal of
    Chemometrics*, 17(3):153–165.
-   Roger, J.-M., Chauchard, F., Bellon-Maurel, V. (2003). EPO-PLS external
    parameter orthogonalisation of PLS application to temperature-independent
    measurement of sugar content of intact fruits. *Chemometrics and
    Intelligent Laboratory Systems*, 66(2):191–204.
-   Trygg, J., Wold, S. (2002). Orthogonal projections to latent structures
    (O-PLS). *Journal of Chemometrics*, 16(3):119–128.
-   Westerhuis, J.A., de Jong, S., Smilde, A.K. (2001). Direct orthogonal
    signal correction. *Chemometrics and Intelligent Laboratory Systems*,
    56(1):13–25.
-   Wold, S., Antti, H., Lindgren, F., Öhman, J. (1998). Orthogonal signal
    correction of near-infrared spectra. *Chemometrics and Intelligent
    Laboratory Systems*, 44(1):175–185.
