---
title: "Removing unwanted variation: a comparison of orthogonalization methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Removing unwanted variation: a comparison of orthogonalization methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  dpi = 96
)
```

Most of the variation in a set of LIBS spectra has nothing to do with the
property we want to predict. Plasma conditions change the total emitted
intensity from shot to shot and sample to sample, and every element in the
sample adds its own lines. Orthogonalization methods try to identify this
unwanted variation and remove it before calibration.

specProc implements two families of such methods, which differ in how
they define "unwanted". Each is available as a function and as a
[recipes](https://recipes.tidymodels.org) step:

| Family | Unwanted variation is… | Recipe steps |
|:-------|:-----------------------|:----------|
| External | the variation between spectra that should be identical (clutter) | `step_epo()`, `step_glsw()` |
| Response-orthogonal | the variation in $\mathbf{X}$ uncorrelated with the response $\mathbf{y}$ | `step_osc()`, `step_direct_orthogonal()`, `step_direct_osc()`, `step_projected_osc()`, `step_y_gradient_glsw()` |

This vignette applies all of them to the same problem, predicting the
potassium content of forage samples, and asks three questions:

1.  **What does each method remove?**
2.  **Does removing it improve predictions** on samples not used to fit
    the filter?
3.  **Which methods are actually the same method?**

Each method is a step in a recipe, combined with a PLS model in a
tidymodels workflow. The workflow re-estimates the filter on the analysis
set of every resample, so supervised filters never see the samples they
are validated on, and the filter is tuned together with the model.

```{r requirements, include = FALSE}
needed <- c("recipes", "parsnip", "workflows", "tune", "rsample", "yardstick",
            "dials", "plsmod", "mixOmics", "purrr")
if (!all(vapply(needed, requireNamespace, logical(1), quietly = TRUE))) {
  knitr::knit_exit(paste("This vignette needs the packages", paste(needed, collapse = ", "),
                         "(mixOmics is on Bioconductor)."))
}
```

```{r setup, message = FALSE}
# parsnip attaches the mixOmics engine, which attaches MASS. Attaching it
# first keeps dplyr::select() (attached with recipes) ahead of MASS::select()
suppressPackageStartupMessages(library(mixOmics))
library(specProc)
library(recipes)
library(parsnip)
library(workflows)
library(tune)
library(rsample)
library(yardstick)
library(plsmod)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)

data(forageLIBS)
elements <- c("Ca", "Cl", "Cu", "Fe", "Mg", "Mn", "Mo", "P", "K", "Na", "S", "Zn")
raw_channels <- setdiff(names(forageLIBS), c("Measurement", "Sample", elements))
wl <- as.numeric(raw_channels)
dim(forageLIBS)
```

`forageLIBS` contains 368 measurements of 365 forage samples (three samples
were measured twice). Each measurement is the mean of 8 laser shots. The
target is potassium, which is available for every measurement and spans
`r round(min(forageLIBS$K), 1)` to `r round(max(forageLIBS$K), 1)`%.

## Preparing the spectra

Two practical steps come first:

-   **Channel overlap.** Two spectrometers overlap near 766 nm, so a few
    channels repeat wavelengths already covered. We drop them.
-   **Binning.** The GLSW filters are $p \times p$ matrices, and
    `glsw()` returns them as such: with 7152 channels, each would take
    about 400 MB. The recipe steps store them in factored form, but
    summing 4 adjacent channels still speeds up every method, at the cost
    of some spectral resolution. Most emission lines in these spectra span
    several channels, so little information is lost.

Finally, `step_snv()` normalizes each spectrum. It works on each spectrum
alone, so it can be applied once to all the data.

```{r bin}
keep <- wl > cummax(c(-Inf, wl[-length(wl)]))
sum(!keep)  # overlapping channels dropped
bin <- ceiling(seq_len(sum(keep)) / 4)
wlb <- as.vector(tapply(wl[keep], bin, mean))
channels <- sprintf("%.3f", wlb)

binned <- as.matrix(forageLIBS[raw_channels[keep]]) |>
  t() |> rowsum(bin) |> t() |>
  as_tibble(.name_repair = \(nm) channels)

unprocessed <- forageLIBS |>
  select(Sample, all_of(elements)) |>
  mutate(total = rowSums(forageLIBS[raw_channels])) |>
  bind_cols(binned)

spectra <- recipe(K ~ ., data = unprocessed) |>
  update_role(Sample, total, all_of(setdiff(elements, "K")), new_role = "id") |>
  step_snv(all_predictors()) |>
  prep() |>
  bake(new_data = NULL)
dim(spectra)
```

## Identifying the unwanted variation

A principal component analysis shows where the variance lies. We correlate
the scores of the first five components with the total emitted intensity
and with the reference contents of several elements:

```{r pca}
covariates <- spectra |> select(total, K, Ca, Mg, Na, P, S)

pca_summary <- function(x, label) {
  pc <- prcomp(x)
  list(
    variance = tibble(data = label, component = paste0("PC", 1:5),
                      percent = round(100 * pc$sdev[1:5]^2 / sum(pc$sdev^2), 1)),
    correlations = cor(pc$x[, 1:5], covariates, use = "pairwise.complete.obs") |>
      round(2) |>
      as_tibble(rownames = "component") |>
      mutate(data = label, .before = 1)
  )
}
raw_pca <- pca_summary(as.matrix(binned), "raw")
snv_pca <- pca_summary(as.matrix(spectra[channels]), "SNV")

bind_rows(raw_pca$variance, snv_pca$variance) |>
  pivot_wider(names_from = component, values_from = percent)   # % variance
raw_pca$correlations
snv_pca$correlations
```

The raw spectra are dominated by one structure. The first component
explains `r raw_pca$variance$percent[1]`% of the variance, and its scores
are almost perfectly correlated with the total intensity. This is the
multiplicative effect of the plasma: a hotter or denser plasma raises all
lines together. SNV removes most of it, but the SNV spectra still carry
several structures that have little to do with potassium:

-   **PC1 and PC2** still follow the total intensity, and also sodium and
    magnesium. After SNV, intensity changes appear as changes in line
    ratios, which a per-spectrum scaling cannot remove.
-   **PC3** contrasts potassium with calcium, the main cation that varies
    in the opposite direction.

How much of the SNV variance is related to potassium at all? Regressing
each channel on $\mathbf{y}$ gives the share of variance that a linear
function of potassium content explains:

```{r y-share}
x_centered <- scale(as.matrix(spectra[channels]), scale = FALSE)
k_centered <- spectra$K - mean(spectra$K)
y_part <- k_centered %*% crossprod(k_centered, x_centered) / sum(k_centered^2)
y_share <- 100 * sum(y_part^2) / sum(x_centered^2)
round(y_share, 1)  # % of the variance
```

Only about `r round(y_share)`% of the variance is directly related to
potassium. The remaining variance is the target of the orthogonalization
methods.

## Calibration and test sets

We set aside one third of the samples as a test set. Filters and PLS models
are always fitted on the calibration set only, and every tuning decision
uses cross-validation within the calibration set, with folds formed by
sample so that the two measurements of a sample stay together.

The external methods need clutter: spectra that should be identical but
are not. The three samples measured twice provide this. Their difference
spectra describe how the measurement of a given material varies from one
session to the next. We keep these three samples in the calibration set,
so the test split is made by hand with `rsample::make_splits()`:

```{r split}
set.seed(1)
twice <- spectra |> count(Sample) |> filter(n > 1) |> pull(Sample)
test_samples <- sample(setdiff(unique(spectra$Sample), twice),
                       round(n_distinct(spectra$Sample) / 3))
is_test <- spectra$Sample %in% test_samples
split <- make_splits(list(analysis = which(!is_test), assessment = which(is_test)),
                     data = spectra)
calibration <- analysis(split)
test <- assessment(split)
c(calibration = nrow(calibration), test = nrow(test))

set.seed(2)
folds <- group_vfold_cv(calibration, group = Sample, v = 5)

# clutter: difference spectra of the samples measured twice
clutter <- spectra |>
  filter(Sample %in% twice) |>
  group_by(Sample) |>
  summarise(across(all_of(channels), \(v) last(v) - first(v))) |>
  select(-Sample)
```

Three difference spectra are a small clutter matrix. In practice, repeated
measurements of a few reference materials over several sessions would
describe this variation better.

## The recipes

Every method is a recipe step added to the same base recipe. The filter
settings are tuned: the number of components removed, or, for GLSW, the
strength $\alpha$ of the down-weighting, relative to the largest
eigenvalue of the clutter. Because the PLS model also has a `num_comp`
parameter, the filter's is named `filter`.

```{r recipes}
base_recipe <- recipe(K ~ ., data = calibration) |>
  update_role(Sample, total, all_of(setdiff(elements, "K")), new_role = "id")

filter_steps <- list(
  `none` = identity,
  # External: the clutter directions are projected out ...
  `EPO` = \(rec, k) step_epo(rec, all_predictors(), clutter = clutter, num_comp = k),
  # ... or down-weighted
  `GLSW` = \(rec, k) step_glsw(rec, all_predictors(), clutter = clutter, alpha = k),
  # Response-orthogonal
  `OSC (Wold)` = \(rec, k) step_osc(rec, all_predictors(), method = "wold", num_comp = k),
  `OSC (Sjoblom)` = \(rec, k) step_osc(rec, all_predictors(), method = "sjoblom", num_comp = k),
  `OSC (Fearn)` = \(rec, k) step_osc(rec, all_predictors(), method = "fearn", num_comp = k),
  `DO` = \(rec, k) step_direct_orthogonal(rec, all_predictors(), num_comp = k),
  `DOSC` = \(rec, k) step_direct_osc(rec, all_predictors(), num_comp = k),
  `POSC / OPLS` = \(rec, k) step_projected_osc(rec, all_predictors(), num_comp = k),
  `y-gradient GLSW` = \(rec, k) step_y_gradient_glsw(rec, all_predictors(), alpha = k)
)
filter_grid <- list(
  `none` = NA, EPO = 1:3, GLSW = 10^-(0:4), `OSC (Wold)` = 1:4, `OSC (Sjoblom)` = 1:4,
  `OSC (Fearn)` = 1:4, DO = 1:4, DOSC = 1:4, `POSC / OPLS` = 1:4, `y-gradient GLSW` = 10^-(0:4)
)

method_recipe <- function(method, k = tune("filter")) {
  if (method == "none") base_recipe else filter_steps[[method]](base_recipe, k)
}
```

`step_opls()` removes the orthogonal components of an `opls()` model. It
gives the same filtered data as `step_projected_osc()`, as shown below, and
is not repeated in the comparison.

## What does each method remove?

We prep each recipe on the calibration set with a comparable setting (2
components, or $\alpha = 0.01$ for GLSW). For the removed part of the
spectra, $\mathbf{X} - f(\mathbf{X})$, we compute:

-   **`removed`:** its share of the calibration variance, in percent.
-   **The correlations of its first principal component** with the total
    intensity and with the element contents (absolute values).

```{r what-removed}
x_cal <- as.matrix(calibration[channels])
settings <- c(EPO = 2, GLSW = 0.01, `OSC (Wold)` = 2, `OSC (Sjoblom)` = 2, `OSC (Fearn)` = 2,
              DO = 2, DOSC = 2, `POSC / OPLS` = 2, `y-gradient GLSW` = 0.01)

removed_parts <- imap(settings, \(k, method) {
  corrected <- method_recipe(method, k) |> prep() |> bake(new_data = NULL)
  scale(x_cal - as.matrix(corrected[channels]), scale = FALSE)
})

removed <- imap(removed_parts, \(r, method) {
  s <- svd(r, nu = 1, nv = 1)
  cor(s$u, select(calibration, total, K, Ca, Mg, Na, P, S), use = "pairwise.complete.obs") |>
    abs() |>
    as_tibble() |>
    mutate(method = method,
           removed = 100 * sum(r^2) / sum(scale(x_cal, scale = FALSE)^2),
           .before = 1)
}) |>
  bind_rows() |>
  mutate(across(where(is.numeric), \(v) round(v, 2)))
removed
```

```{r removed-numbers, include = FALSE}
pick <- function(method, column) removed[[column]][removed$method == method]
y_orth <- c("OSC (Wold)", "OSC (Sjoblom)", "OSC (Fearn)", "DO", "DOSC", "POSC / OPLS")
```

The table shows what each family targets:

-   **The response-orthogonal methods remove the most variance**, about
    `r round(min(sapply(y_orth, pick, "removed")))`–`r round(max(sapply(y_orth, pick, "removed")))`%
    with only two components. The removed part is uncorrelated with
    potassium, exactly so for Wold's and Fearn's OSC, DOSC and POSC. DO
    and Sjöblom's OSC keep a small correlation, because they remove the
    principal directions of the $\mathbf{y}$-orthogonal space, and the
    scores of $\mathbf{X}$ on these directions need not be orthogonal to
    $\mathbf{y}$. What these methods remove is strongly related to the
    total intensity and to sodium, the structures found by the PCA above.
-   **EPO and GLSW remove less**, and what they remove is mostly the
    total intensity (correlations of `r pick("GLSW", "total")` and
    `r pick("EPO", "total")`). When a sample is measured again, what
    changes most is the plasma, not the composition. It also shows that
    SNV does not remove all intensity effects.
-   **y-gradient GLSW** removes about
    `r round(pick("y-gradient GLSW", "removed"))`% of the variance at
    $\alpha = 0.01$, again related to sodium and intensity. It
    down-weights directions rather than removing them, so $\alpha$ sets its
    strength continuously, from almost nothing to a projection like EPO.

The first direction of the removed part (the loading) shows which
emission lines are involved:

```{r directions, fig.height = 7}
lines_nm <- c(Mg = 279.55, Ca = 393.37, Na = 588.99, H = 656.28, K = 766.49, O = 777.19)

removed_parts[c("EPO", "DO", "POSC / OPLS", "y-gradient GLSW")] |>
  imap(\(r, method) tibble(method = method, wavelength = wlb,
                           loading = svd(r, nu = 0, nv = 1)$v[, 1])) |>
  bind_rows() |>
  mutate(method = factor(method, levels = c("EPO", "DO", "POSC / OPLS", "y-gradient GLSW"))) |>
  ggplot(aes(wavelength, loading)) +
  geom_vline(xintercept = lines_nm, colour = "grey60", linetype = "dotted") +
  geom_line(linewidth = 0.3) +
  geom_text(data = tibble(wavelength = lines_nm, label = names(lines_nm)),
            aes(wavelength, Inf, label = label), vjust = 1.2, size = 2.5, inherit.aes = FALSE) +
  facet_wrap(~ method, ncol = 1, scales = "free_y") +
  labs(x = "Wavelength (nm)", y = "Loading") +
  theme_bw()
```

The sign of each loading is arbitrary. The four directions share a common
pattern: the emission lines of the sample's metals (Mg, Ca and above all
the Na doublet at 589 nm) vary against the O I line at 777 nm and the
continuum above 800 nm. For a measurement in air, the O I line comes from
both the atmosphere and the organic matrix, so this contrast likely
reflects how much material is ablated and how the plasma develops rather
than the composition of the sample. The sodium doublet is the largest
feature for the response-orthogonal methods and y-gradient GLSW: its
variation between samples is large and independent of potassium. The
potassium lines barely appear in these directions.

## Does removing it improve predictions?

We now combine each recipe with a PLS model and tune both the filter
parameter and the number of PLS components (up to 15) with
`tune_grid()` on the grouped folds of the calibration set:

```{r tuning}
# mixOmics scales every channel by default; scale = FALSE only centers
pls_model <- pls(num_comp = tune()) |>
  set_mode("regression") |>
  set_engine("mixOmics", scale = FALSE)

tune_method <- function(method) {
  grid <- tibble(num_comp = 1:15)
  if (method != "none") grid <- expand_grid(filter = filter_grid[[method]], grid)
  suppressMessages(tune_grid(workflow(method_recipe(method), pls_model),
                             resamples = folds, grid = grid, metrics = metric_set(rmse)))
}
tuned <- map(set_names(names(filter_steps)), tune_method)
```

The best setting of each method is then refitted to the whole
calibration set and evaluated on the test set with `last_fit()`:

```{r compare}
final_fits <- imap(tuned, \(res, method) {
  workflow(method_recipe(method), pls_model) |>
    finalize_workflow(select_best(res, metric = "rmse")) |>
    last_fit(split, metrics = metric_set(rmse))
})

settings_chosen <- imap(tuned, \(res, method) {
  best <- show_best(res, metric = "rmse", n = 1)
  tibble(method = method,
         filter = if (method == "none") NA_real_ else best$filter,
         pls_ncomp = best$num_comp,
         RMSECV = best$mean)
}) |>
  bind_rows()

test_predictions <- final_fits |>
  map(\(f) collect_predictions(f) |> select(.row, K, .pred)) |>
  bind_rows(.id = "method") |>
  pivot_wider(names_from = method, values_from = .pred)
```

With `r nrow(test)` test spectra, the test error itself is uncertain.
Because all methods predict the same test samples, we compare each one
with plain PLS (`none`) using a paired bootstrap of the test samples.
`diff_low` and `diff_high` bound a 95% interval for the difference in
RMSEP; negative values favor the filter.

```{r compare-table}
methods <- names(filter_steps)
rmse_of <- \(d) summarise(d, across(all_of(methods), \(p) sqrt(mean((p - K)^2))))

set.seed(3)
boot_diff <- bootstraps(test_predictions, times = 2000) |>
  mutate(rmse = map(splits, \(s) rmse_of(analysis(s)))) |>
  select(rmse) |>
  unnest(rmse) |>
  mutate(across(all_of(methods), \(v) v - none))

comparison <- settings_chosen |>
  left_join(rmse_of(test_predictions) |> pivot_longer(everything(), names_to = "method",
                                                      values_to = "RMSEP"), by = "method") |>
  left_join(boot_diff |>
              pivot_longer(everything(), names_to = "method") |>
              group_by(method) |>
              summarise(diff_low = quantile(value, 0.025), diff_high = quantile(value, 0.975)),
            by = "method") |>
  mutate(across(c(RMSECV, RMSEP, diff_low, diff_high), \(v) round(v, 3)))
comparison
c(null_RMSEP = round(sqrt(mean((mean(calibration$K) - test$K)^2)), 3))
```

```{r compare-numbers, include = FALSE}
row_of <- \(m) comparison[comparison$method == m, ]
identical_to_pls <- comparison$method[comparison$method != "none" &
                                        comparison$diff_low == 0 & comparison$diff_high == 0]
better <- comparison$method[comparison$diff_high < 0]
fewest <- comparison |> filter(method != "none") |> slice_min(pls_ncomp, n = 1, with_ties = FALSE)
```

All models predict potassium far better than the null model, which
predicts the calibration mean for every sample. The comparison with plain
PLS:

-   **No filter improves predictions measurably**
    `r if (length(better) == 0) "." else paste0("except ", paste(better, collapse = ", "), ".")`
    The intervals for the difference with plain PLS include zero or favor
    plain PLS, and the differences in RMSEP are a few hundredths of a
    percent of potassium, against a typical error of about
    `r round(row_of("none")$RMSEP, 2)`%.
-   **Filtering mostly moves components from PLS to the filter.** The
    component-based filters remove components and need fewer PLS
    components in return (compare `pls_ncomp` with the
    `r row_of("none")$pls_ncomp` components of plain PLS); the total model
    complexity hardly changes.
-   **Some filtered models are plain PLS in disguise.**
    `r if (length(identical_to_pls) > 0) paste0("The ", paste(identical_to_pls, collapse = " and "), " models give exactly the same test predictions as plain PLS (both bootstrap bounds are 0), as the next section explains.") else "OPLS-type filters followed by PLS reproduce PLS models, as the next section explains."`
-   **The most parsimonious model** is the `r fewest$method` model, with
    `r fewest$pls_ncomp` PLS components, but its RMSEP is not
    distinguishable from that of plain PLS.

The external filters could not help much: their clutter comes from three
samples, and the main thing it describes, the intensity effect, is
variation PLS already learns to ignore from `r nrow(calibration)`
calibration spectra.

## Which methods are the same method?

Several functions compute the same filter, or a filter that cannot change
PLS predictions:

```{r equivalences}
# POSC, O2PLS with a single response and OPLS all give OPLS-filtered data
posc <- as.matrix(projected_osc(x_cal, calibration$K, ncomp = 3)$correction)
all.equal(posc, as.matrix(o2pls(x_cal, calibration$K, ncomp = 1, nx = 2)$correction),
          check.attributes = FALSE)
all.equal(posc, as.matrix(opls(x_cal, calibration$K, ncomp.ortho = 2, permutation = 0)$correction),
          check.attributes = FALSE)

# OPLS filter with 2 orthogonal components + 1-component PLS
# gives the same predictions as a 3-component PLS model
predict_with <- function(rec, ncomp) {
  workflow(rec, set_args(pls_model, num_comp = ncomp)) |>
    fit(data = calibration) |>
    predict(new_data = test) |>
    pull(.pred)
}
all.equal(predict_with(method_recipe("POSC / OPLS", 2), 1),
          predict_with(base_recipe, 3))
```

The first two results mean that `projected_osc()`, `o2pls()` (with one
response) and `opls()` are interchangeable as filters (and so are their
steps). The third explains why OPLS-filtered
models can reproduce plain PLS predictions exactly. OPLS splits the
systematic variation of a PLS model into a predictive and an orthogonal
part, but the model is the same (Trygg and Wold, 2002; Kemsley and Tapp,
2009). These filters are still useful for interpretation: the orthogonal
components show what varies in the spectra independently of potassium,
as in the figure above.

## The main pitfall: fitting the filter before validating

The response-orthogonal filters use $\mathbf{y}$. If a filter is fitted
once on all calibration samples and cross-validation is run afterwards,
the validation samples have already shaped the filter. We prep a DOSC
recipe on the whole calibration set, bake the data, and only then
cross-validate a PLS model on the filtered spectra:

```{r leak}
leak <- map(c(1, 2, 4, 8), \(k) {
  filtered_recipe <- method_recipe("DOSC", k) |> prep()
  filtered_cal <- bake(filtered_recipe, new_data = NULL)
  filtered_folds <- manual_rset(
    map(folds$splits, \(s) make_splits(list(analysis = s$in_id, assessment = complement(s)),
                                       data = filtered_cal)),
    folds$id)
  cv <- tune_grid(workflow(base_recipe, pls_model), resamples = filtered_folds,
                  grid = tibble(num_comp = 1:15), metrics = metric_set(rmse))
  best <- select_best(cv, metric = "rmse")
  test_pred <- finalize_workflow(workflow(base_recipe, pls_model), best) |>
    fit(data = filtered_cal) |>
    predict(new_data = bake(filtered_recipe, new_data = test)) |>
    pull(.pred)
  tibble(filter = k, pls_ncomp = best$num_comp,
         leaky_RMSECV = show_best(cv, metric = "rmse", n = 1)$mean,
         RMSEP = sqrt(mean((test_pred - test$K)^2)))
}) |>
  bind_rows() |>
  mutate(across(c(leaky_RMSECV, RMSEP), \(v) round(v, 3)))
leak
```

With the filter fitted before cross-validation, the cross-validated error
is close to zero, while the error on the test set is higher than for plain
PLS. DOSC first projects $\mathbf{y}$ onto the space of $\mathbf{X}$. With
far more channels than samples, this projection reproduces the calibration
values almost exactly, and so does the filtered $\mathbf{X}$. The
calibration vignette (`vignette("calibration", package = "specProc")`)
shows a milder version of the same bias for the OPLS filter. In the
comparison above, the filter is a step of the workflow's recipe, so
`tune_grid()` refits it inside every fold and this bias cannot occur.

## How much of the signal serves the prediction?

The filters above remove variation; the net analyte signal (NAS) asks the
complementary question. The NAS of a spectrum is the part that is unique to
the analyte, orthogonal to the contributions of all other constituents.
For an inverse model such as PLS, it follows from the regression vector
$\mathbf{b}$ (Lorber *et al.*, 1997; Faber, 1998): the NAS of a sample is
$(\hat{y}_i - \bar{y}) / \|\mathbf{b}\|$. `nas()` computes it together with
the figures of merit of the model (Olivieri *et al.*, 2006):

-   **Sensitivity**, $1 / \|\mathbf{b}\|$: the NAS produced by 1% of
    potassium, in SNV units.
-   **Selectivity**: the fraction of each (centered) spectrum that is used
    for prediction.
-   **Limit of detection and of quantification**, $3.3\,\sigma / \mathrm{SEN}$
    and $10\,\sigma / \mathrm{SEN}$, where $\sigma$ is the noise level of the
    spectra.

We estimate $\sigma$ from the three samples measured twice: the standard
deviation of their difference spectra, divided by $\sqrt{2}$. This is a
repeatability between measurements, which includes the plasma effects seen
above, not the noise of the detector alone.

```{r nas}
noise <- sqrt(mean(as.matrix(clutter)^2) / 2)
merit <- map(c(`5 components` = 5, `10 components` = 10), \(a) {
  nas(x_cal, calibration$K, ncomp = a, noise = noise)$figures_of_merit
}) |>
  as_tibble() |>
  mutate(figure = names(nas(x_cal, calibration$K, ncomp = 5, noise = noise)$figures_of_merit),
         .before = 1) |>
  mutate(across(-figure, \(v) round(v, 3)))
merit

# selectivity of the test samples, 10-component model
fit_nas <- nas(x_cal, calibration$K, ncomp = 10, noise = noise)
summary(predict(fit_nas, test[channels])$selectivity)
```

```{r nas-numbers, include = FALSE}
fom <- \(figure, model) merit[[model]][merit$figure == figure]
```

Two results stand out:

-   **Only a small part of each spectrum carries potassium information.**
    With 10 components, the mean selectivity is
    `r fom("selectivity", "10 components")`: less than a tenth of the
    centered signal is used for prediction, the rest is shared with
    other constituents or is plasma variation. This is the same conclusion
    as the variance share computed at the start, seen from the model's side.
-   **More components trade sensitivity for selectivity.** Going from 5 to
    10 components more than halves the sensitivity, because the regression
    vector must avoid more interferent directions, and more than doubles
    the limit of detection, to `r round(fom("lod", "10 components"), 2)`% K.
    The 10-component model still predicts better: its lower sensitivity is
    the price of being less affected by the interferents.

These limits account for the repeatability of the spectra only, not for
the errors of the reference analyses or of the model itself, so they are
lower bounds. Before specProc 0.4.0, `nas()` returned spectra filtered by
direct orthogonalization; `direct_orthogonal()` computes that filter.

## Summary

-   **Identify first.** A PCA with correlations to known factors (total
    intensity, other elements) shows what the unwanted variation is. Here,
    it is mainly the plasma intensity effect and the variation of sodium
    and calcium.
-   **External methods remove what you tell them to.** EPO and GLSW are
    only as good as the clutter. Differences between repeated measurements
    of the same material captured the intensity effect, which SNV leaves
    in part.
-   **Response-orthogonal methods remove a lot of variance, but add no
    information.** They make models smaller and easier to interpret, not
    more accurate: the removed variation is variation that PLS already
    ignores. Some (OPLS/POSC, O2PLS, Fearn's OSC) can give exactly the
    PLS predictions.
-   **Validate honestly.** Put supervised filters in the recipe of a
    workflow, so they are refitted inside each fold. Otherwise the
    cross-validated error can be meaningless.

## References

-   Andersson, C.A. (1999). Direct orthogonalization. *Chemometrics and
    Intelligent Laboratory Systems*, 47(1):51–63.
-   Fearn, T. (2000). On orthogonal signal correction. *Chemometrics and
    Intelligent Laboratory Systems*, 50(1):47–52.
-   Faber, N.M. (1998). Efficient computation of net analyte signal vector
    in inverse multivariate calibration models. *Analytical Chemistry*,
    70(23):5108–5110.
-   Kemsley, E.K., Tapp, H.S. (2009). OPLS filtered data can be obtained
    directly from non-orthogonalized PLS1. *Journal of Chemometrics*,
    23(5):263–264.
-   Lorber, A., Faber, K., Kowalski, B.R. (1997). Net analyte signal
    calculation in multivariate calibration. *Analytical Chemistry*,
    69(8):1620–1626.
-   Martens, H., Høy, M., Wise, B.M., Bro, R., Brockhoff, P.B. (2003).
    Pre-whitening of data by covariance-weighted preprocessing. *Journal of
    Chemometrics*, 17(3):153–165.
-   Olivieri, A.C., Faber, N.M., Ferré, J., Boqué, R., Kalivas, J.H.,
    Mark, H. (2006). Uncertainty estimation and figures of merit for
    multivariate calibration (IUPAC Technical Report). *Pure and Applied
    Chemistry*, 78(3):633–661.
-   Roger, J.-M., Chauchard, F., Bellon-Maurel, V. (2003). EPO-PLS external
    parameter orthogonalisation of PLS application to temperature-independent
    measurement of sugar content of intact fruits. *Chemometrics and
    Intelligent Laboratory Systems*, 66(2):191–204.
-   Trygg, J., Wold, S. (2002). Orthogonal projections to latent structures
    (O-PLS). *Journal of Chemometrics*, 16(3):119–128.
-   Westerhuis, J.A., de Jong, S., Smilde, A.K. (2001). Direct orthogonal
    signal correction. *Chemometrics and Intelligent Laboratory Systems*,
    56(1):13–25.
-   Wold, S., Antti, H., Lindgren, F., Öhman, J. (1998). Orthogonal signal
    correction of near-infrared spectra. *Chemometrics and Intelligent
    Laboratory Systems*, 44(1):175–185.
