---
title: "Plasma diagnostics: electron density, temperature and self-absorption"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plasma diagnostics: electron density, temperature and self-absorption}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  dpi = 96
)
```

Quantitative LIBS relies on the plasma being in local thermodynamic
equilibrium (LTE), optically thin, and recorded within the linear range of
the detector. This vignette shows how to check these conditions and how
to measure the two quantities that describe the plasma: the electron
density \(N_e\), from the Stark broadening of a line, and the excitation
temperature \(T\), from Boltzmann and Saha-Boltzmann plots. It uses the
spectra of `specLIBS` and `fourrage`, atomic data from the NIST Atomic
Spectra Database, and Stark broadening parameters from the STARK-B
database.

Two limitations of these data affect every number below:

-   **The intensities are not corrected for the spectral response** of the
    instrument, which weights lines at different wavelengths differently.
    The temperatures are therefore indicative.
-   **The acquisition delay and gate are not known.** \(N_e\) and \(T\) change
    quickly as the plasma cools, and the values are averages over the gate.

```{r setup, message = FALSE}
library(specProc)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

data(specLIBS)
meta_cols <- c("Sample", "Location", "Clay", "Sand", "Silt", "Texture", "Structure", "Type")
channels <- setdiff(names(specLIBS), meta_cols)
wl <- as.numeric(channels)
```

## Detector saturation

A saturated line is clipped at the detector's maximum count: its area and
width are wrong, and it no longer grows with concentration. It must be
excluded from any diagnostic. `saturation_summary()` lists the saturated
channels. The `fourrage` spectra are means of 8 shots, so we count as
saturated the channels whose mean is within 1% of the 16-bit limit (65535),
that is, saturated in nearly every shot:

```{r saturation}
data(fourrage)
saturated <- saturation_summary(fourrage[-(1:14)], limit = 65535, tolerance = 655)

# adjacent saturated channels belong to the same line
saturated$channels |>
  mutate(line = cumsum(c(1, diff(index) > 2))) |>
  group_by(line) |>
  summarise(wavelength = round(mean(wavelength), 2), channels = n(),
            measurements = max(n_spectra)) |>
  select(-line) |>
  arrange(desc(measurements))
```

The strongest lines of the forage spectra are saturated in a large share
of the measurements: Ca II 393.37 and 396.85 nm, Mg II 279.55 and 280.27
nm, the Ca I resonance line at 422.67 nm, and some Na and K lines. None of
them can be used for plasma diagnostics in these data. The `specLIBS`
spectra peak at `r max(specLIBS[channels])` counts, well below the limit:

```{r saturation-libs}
nrow(saturation_summary(specLIBS[channels], limit = 65535)$channels)
```

We use `specLIBS` below, averaged over all 400 spectra after baseline
correction, which gives a single high signal-to-noise spectrum:

```{r mean-spectrum}
spectrum <- baseline_arpls(specLIBS[channels], lambda = 1e5, max.iter = 20)$correction |>
  colMeans()
```

## Electron density from the H\(\alpha\) line

Water and organic matter in the soil give a hydrogen line at 656.28 nm.
H\(\alpha\) is the standard density diagnostic in LIBS: its Stark width is
large, it is rarely self-absorbed, and computer simulations relate its
width to \(N_e\) with little dependence on temperature (Gigosos et al.,
2003). The relation between width and density applies to the full width
at half maximum (FWHM) of H\(\alpha\)'s Stark profile, which is not
Lorentzian. We therefore fit a Voigt profile only to describe the line
shape, and use its total FWHM:

```{r halpha}
halpha_window <- channels[wl > 654.6 & wl < 658.3]
halpha_fit <- as_tibble(as.list(spectrum[halpha_window])) |>
  peak_fit(profile = "voigt")
halpha_fit$tidied[[1]]
```

```{r halpha-plot, fig.height = 5}
plot_fit(halpha_fit, title = "H-alpha line, mean of all specLIBS spectra")
```

```{r halpha-density}
widths <- halpha_fit$tidied[[1]] |> select(term, estimate) |> deframe()
fwhm_halpha <- voigt_fwhm(widths[["wG"]], widths[["wL"]])
ne <- electron_density(fwhm_halpha, method = "halpha")
c(fwhm_nm = fwhm_halpha, electron_density = ne)
```

The Gaussian and Lorentzian widths of the fit have large standard errors:
with a noisy line, the fit can trade one for the other. Their combination,
the total width, is much better determined. The instrumental width, about
0.1 to 0.2 nm judging from the narrow lines of these spectra, is small
compared with the FWHM of `r round(fwhm_halpha, 2)` nm and changes it by
less than 1% when subtracted in quadrature, so we neglect it. The width
gives \(N_e \approx\) `r format(signif(ne, 2))` cm\(^{-3}\), a typical
value for LIBS plasmas in air a few hundred nanoseconds to a few
microseconds after the laser pulse.

## Temperature from Boltzmann plots

### Atomic data

A Boltzmann plot needs, for each line, the transition probability
\(A_{ki}\), the statistical weight \(g_k\) and the energy \(E_k\) of the upper
level. `nist_lines()` retrieves them from the NIST Atomic Spectra
Database, and `nist_ionization_energy()` the ionization energies:

```{r nist-fetch, eval = FALSE}
ca_neutral <- nist_lines("Ca I", wavelength = c(420, 660))
ca_ion <- nist_lines("Ca II", wavelength = c(310, 400))
ionization_energy <- nist_ionization_energy("Ca I")
```

So that this vignette runs without an internet connection, the values
retrieved for the calcium lines used below are included here (NIST ASD,
Kramida et al.):

```{r atomic-data}
atomic <- tribble(
  ~species, ~stage, ~wavelength, ~Aki,    ~accuracy, ~Ek,       ~gk,
  "Ca I",   1,      428.301,     4.34e7,  "C+",      4.779784,  5,
  "Ca I",   1,      430.253,     1.36e8,  "C+",      4.779784,  5,
  "Ca I",   1,      431.865,     7.40e7,  "C+",      4.7690284, 3,
  "Ca I",   1,      443.496,     6.70e7,  "C",       4.680635,  5,
  "Ca I",   1,      445.478,     8.70e7,  "C",       4.681327,  7,
  "Ca I",   1,      558.876,     4.90e7,  "D",       4.7435268, 7,
  "Ca I",   1,      610.272,     9.60e6,  "C",       3.910399,  3,
  "Ca I",   1,      612.222,     2.87e7,  "C",       3.910399,  3,
  "Ca I",   1,      616.217,     4.77e7,  "C",       3.910399,  3,
  "Ca I",   1,      643.907,     5.30e7,  "D",       4.450647,  9,
  "Ca I",   1,      646.257,     4.70e7,  "D",       4.4409544, 7,
  "Ca I",   1,      649.378,     4.40e7,  "D",       4.4300117, 5,
  "Ca II",  2,      315.887,     3.10e8,  "C",       7.047169,  4,
  "Ca II",  2,      317.933,     3.60e8,  "C",       7.049551,  6,
  "Ca II",  2,      370.603,     8.80e7,  "C",       6.467875,  2,
  "Ca II",  2,      373.690,     1.70e8,  "C",       6.467875,  2
)
ionization_energy <- 6.1131549   # Ca I, eV (NIST ASD)
```

The accuracy grades of these transition probabilities (C: ≤ 25%, D:
≤ 50%) limit the precision of any temperature derived from them. The Ca II
resonance lines at 393.37 and 396.85 nm are left out: they are strongly
self-absorbed, as shown at the end of this vignette.

### Line intensities

The spectrometer's wavelength scale is offset from the NIST wavelengths by
about one channel in places, so each line area is summed over ±0.15 nm
around the observed peak within 0.2 nm of the tabulated wavelength:

```{r line-areas}
line_area <- function(center, search = 0.2, half_width = 0.15) {
  near <- abs(wl - center) <= search
  peak <- wl[near][which.max(spectrum[near])]
  sum(spectrum[abs(wl - peak) <= half_width])
}
lines <- atomic |> mutate(intensity = sapply(wavelength, line_area))
```

### Boltzmann plot of the neutral atom

```{r boltzmann}
neutral <- boltzmann_plot(filter(lines, stage == 1))
neutral
```

```{r boltzmann-plot, fig.height = 4}
plot_boltzmann(neutral)
```

### Saha-Boltzmann plot

Adding lines of the ion extends the energy range. The Saha equation places
them on the same plot, using the electron density measured on
H\(\alpha\):

```{r saha}
saha <- saha_boltzmann_plot(lines, ionization_energy = ionization_energy, electron_density = ne)
saha
```

```{r saha-plot, fig.height = 4}
plot_boltzmann(saha)
```

The Ca I lines come from upper levels between 3.9 and 4.8 eV. Over such a
narrow range, the slope of the Boltzmann plot is sensitive to errors in a
few intensities or transition probabilities (graded C and D, that is,
uncertain by 25 to 50%): its temperature,
`r round(neutral$temperature, -1)` ± `r round(neutral$temperature_se, -1)`
K, is far below the Saha-Boltzmann estimate. This is the usual problem of
Boltzmann plots with lines of a single species.

The ionic points extend the abscissa by the ionization energy, to about 13
eV, and the Saha-Boltzmann fit gives \(T \approx\)
`r round(saha$temperature, -2)` K with a standard error of
`r round(saha$temperature_se, -1)` K. The standard error reflects the
scatter of the points, not the systematic errors of the transition
probabilities, the spectral response, or the density.

### Is the plasma in LTE?

The McWhirter criterion gives the minimum density for collisions to
dominate the population of the levels. The largest energy gap here is the
first excited level of Ca II, 3.15 eV above the ground state:

```{r mcwhirter}
lte <- mcwhirter_criterion(temperature = saha$temperature, delta_e = 3.15, electron_density = ne)
lte
```

The measured density is `r round(ne / lte$minimum_density)` times the
minimum. The criterion is necessary but not sufficient: in a transient,
inhomogeneous plasma, LTE also requires equilibration to be faster than
the changes of the plasma (Cristoforetti et al., 2010).

## Stark broadening parameters from STARK-B

For lines other than H\(\alpha\), the electron density follows from the
Stark width in the impact approximation, \(N_e = N_{ref}\,w / w_{ref}(T)\),
with \(w_{ref}\) from STARK-B. `starkb_lines()` downloads the widths and
shifts of a species:

```{r starkb, eval = requireNamespace("xml2", quietly = TRUE)}
stark <- tryCatch(
  starkb_lines("Ca II", wavelength = c(390, 400), perturber = "electron"),
  error = function(e) NULL
)
if (is.null(stark)) {
  message("STARK-B could not be reached; the sections that need it are skipped.")
} else {
  stark |> filter(density == 1e17) |> select(wavelength, upper, lower, temperature, width, shift)
}
```

```{r stark-available, include = FALSE}
has_stark <- exists("stark") && !is.null(stark) && nrow(stark) > 0
```

STARK-B tabulates the Ca II 4s–4p multiplet at its mean wavelength.
`stark_width()` interpolates the width at the plasma temperature, scales
it linearly to the electron density, and converts the multiplet width to
that of the line with the \(\lambda^2\) rule:

```{r stark-width, eval = has_stark}
thin <- stark_width(stark, wavelength = 393.366, temperature = saha$temperature,
                    density = ne, tolerance = 2)
thin
```

If the service is not available, a table saved earlier can be read with
`read_starkb()`, and widths from other sources can be entered with
`stark_table()`, which takes tabulated values or STARK-B's fitted
temperature laws:

```{r stark-table}
# made-up widths of a hypothetical line, in Angstrom at 1e17 cm-3
stark_table(wavelength = 5000, temperature = c(5000, 10000, 20000),
            width = c(0.20, 0.16, 0.13), units = "A", species = "X II")
```

Take the values of your line from the database or the literature, and
state their units, since widths are published in both Å and nm.

## Self-absorption of a resonance line

Resonance lines end on the ground state, which is heavily populated, so
the plasma reabsorbs their light. Self-absorption flattens and broadens
the line: its measured Stark width exceeds the width expected at the
plasma's electron density. The ratio gives the self-absorption
coefficient \(SA\) (El Sherbini et al., 2005), the fraction of the peak
intensity that escapes. Ca II 393.37 nm is an isolated line, so its Stark
profile is Lorentzian: we fit it jointly with its neighbors and take its
Lorentzian width:

```{r ca-fit, eval = has_stark}
ca_window <- channels[wl > 392.9 & wl < 397.3]
ca_fit <- as_tibble(as.list(spectrum[ca_window])) |>
  multipeak_fit(peaks = c(393.37, 394.40, 396.15, 396.85), profiles = "voigt")
measured <- ca_fit$tidied[[1]] |> filter(term == "wL_1")
measured
self_absorption(width = measured$estimate, thin_width = thin$width)
```

```{r sa-numbers, include = FALSE, eval = has_stark}
sa <- self_absorption(width = measured$estimate, thin_width = thin$width)
```

`r if (has_stark) paste0("The measured Lorentzian width, ", round(measured$estimate, 3), " nm, is ", round(measured$estimate / thin$width, 1), " times the optically thin width of ", round(thin$width, 3), " nm, which gives SA of about ", signif(sa$SA, 1), ": only a fraction of the peak intensity escapes the plasma.") else "(This section needs the STARK-B service.)"`
The Lorentzian width of a line only a few channels wide is uncertain (see
its standard error), so this coefficient is an order of magnitude rather
than a correction factor. It confirms what the saturation of the same line
in the forage spectra already suggested: the Ca II resonance lines are
unsuitable for quantification in these plasmas.

## Summary

-   **Screen for saturation first.** Saturated lines are useless for
    diagnostics and nonlinear for calibration.
-   **Measure \(N_e\) on a line that is broad and optically thin,** such as
    H\(\alpha\), from the FWHM of its Stark profile; for isolated lines, use
    the Lorentzian width of a Voigt fit with Stark data from STARK-B.
-   **Prefer Saha-Boltzmann to Boltzmann plots.** Lines of two ionization
    stages span a much wider energy range; check the accuracy grades of the
    transition probabilities, and correct the intensities for the spectral
    response when a calibration is available.
-   **Check LTE and self-absorption** before trusting intensities: the
    McWhirter criterion is a minimum requirement, and resonance lines are
    often strongly self-absorbed.

## References

-   Cristoforetti, G., De Giacomo, A., Dell'Aglio, M., Legnaioli, S.,
    Tognoni, E., Palleschi, V., Omenetto, N. (2010). Local thermodynamic
    equilibrium in laser-induced breakdown spectroscopy: beyond the
    McWhirter criterion. *Spectrochimica Acta Part B*, 65(1):86–95.
-   El Sherbini, A.M., et al. (2005). Evaluation of self-absorption
    coefficients of aluminum emission lines in laser-induced breakdown
    spectroscopy measurements. *Spectrochimica Acta Part B*,
    60(12):1573–1579.
-   Gigosos, M.A., González, M.Á., Cardeñoso, V. (2003). Computer simulated
    Balmer-alpha, -beta and -gamma Stark line profiles for non-equilibrium
    plasmas diagnostics. *Spectrochimica Acta Part B*, 58(8):1489–1504.
-   Kramida, A., Ralchenko, Yu., Reader, J., and NIST ASD Team. NIST Atomic
    Spectra Database, <https://physics.nist.gov/asd>.
-   Olivero, J.J., Longbothum, R.L. (1977). Empirical fits to the Voigt
    line width: a brief review. *Journal of Quantitative Spectroscopy and
    Radiative Transfer*, 17(2):233–236.
-   Sahal-Bréchot, S., Dimitrijević, M.S., Moreau, N. STARK-B database,
    <https://stark-b.obspm.fr>. Observatoire de Paris and Astronomical
    Observatory of Belgrade.
