---
title: "4. Plasma diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Plasma diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  dpi = 96
)
```

Quantitative LIBS assumes that the plasma is in local thermodynamic
equilibrium (LTE), optically thin, and recorded within the linear range of
the detector. This vignette checks these conditions on the forage spectra,
measures the electron density \(N_e\) and the temperature \(T\) of the
plasma, and ends with a calibration-free estimate of the composition of the
samples. The intensities are not corrected for the spectral response of the
instrument, and the acquisition delay is not known, so the values are
indicative.

```{r setup, message = FALSE}
library(specProc)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

data("forageLIBS")
spectra_id <- forageLIBS |> select(1:2) |> names()
minerals <- forageLIBS |> select(3:14) |> names()
spectra <- forageLIBS |> select(-all_of(c(spectra_id, minerals)))
wl <- as.numeric(names(spectra))

corrected <- baseline_arpls(spectra, lambda = 1e5, max.iter = 20)$correction
mean_spectrum <- colMeans(corrected)
```

## Detector saturation

A saturated line is clipped at the maximum count of the detector: its area
is wrong and no longer grows with concentration. `saturation_summary()`
lists the saturated channels. Each spectrum is the mean of 8 shots, so we
count the channels within 1% of the 16-bit limit (65535), that is,
saturated in nearly every shot:

```{r saturation}
saturated <- saturation_summary(spectra, limit = 65535, tolerance = 655)

# adjacent saturated channels belong to the same line
saturated$channels |>
  mutate(line = cumsum(c(1, diff(index) > 2))) |>
  group_by(line) |>
  summarise(wavelength = round(mean(wavelength), 2), spectra = max(n_spectra)) |>
  select(-line) |>
  arrange(desc(spectra))
```

The resonance lines of Ca II (393.37 and 396.85 nm), Mg II (279.55 and
280.27 nm), Ca I (422.67 nm), Na I (589.00 and 589.59 nm) and K I (769.90
nm), as well as the head of the CN molecular band at 388.3 nm, are
saturated in part of the spectra: they are left out of every diagnostic
below.

## Electron density

The hydrogen line H\(\alpha\) (656.28 nm) is the standard density
diagnostic: its Stark width is large and depends little on temperature
(Gigosos et al., 2003). Its profile is Lorentzian in these spectra (see
`vignette("peak-fitting")`), and `electron_density()` converts its full
width at half maximum into \(N_e\):

```{r density}
halpha <- as_tibble(as.list(mean_spectrum[wl > 653.5 & wl < 659.5])) |>
  peak_fit(profile = "lorentzian")
width <- halpha$tidied[[1]] |> filter(term == "wL")
width
ne <- electron_density(width$estimate, method = "halpha")
ne
```

For other lines, `stark_width()` gives the width expected at a density from
the Stark parameters of the STARK-B database (`starkb_lines()`), and
`self_absorption()` compares it with the measured width.

## Temperature

### Atomic data

A Boltzmann plot needs, for each line, the transition probability
\(A_{ki}\), the statistical weight \(g_k\) and the energy \(E_k\) of the upper
level. `nist_lines()` retrieves them from the NIST Atomic Spectra Database,
and `nist_ionization_energy()` the ionization energies:

```{r nist, eval = FALSE}
ca_lines <- nist_lines(c("Ca I", "Ca II"), wavelength = c(310, 660))
nist_ionization_energy("Ca I")
```

So that this vignette runs offline, the calcium lines used below are
included (NIST ASD, Kramida et al.). The accuracy grades of the transition
probabilities (C: ≤ 25%, D: ≤ 50%) limit the precision of the temperature:

```{r atomic}
atomic <- tribble(
  ~species, ~stage, ~wavelength, ~Aki,    ~accuracy, ~Ek,       ~gk,
  "Ca I",   1,      428.301,     4.34e7,  "C+",      4.779784,  5,
  "Ca I",   1,      430.253,     1.36e8,  "C+",      4.779784,  5,
  "Ca I",   1,      431.865,     7.40e7,  "C+",      4.7690284, 3,
  "Ca I",   1,      443.496,     6.70e7,  "C",       4.680635,  5,
  "Ca I",   1,      445.478,     8.70e7,  "C",       4.681327,  7,
  "Ca I",   1,      558.876,     4.90e7,  "D",       4.7435268, 7,
  "Ca I",   1,      610.272,     9.60e6,  "C",       3.910399,  3,
  "Ca I",   1,      612.222,     2.87e7,  "C",       3.910399,  3,
  "Ca I",   1,      616.217,     4.77e7,  "C",       3.910399,  3,
  "Ca I",   1,      643.907,     5.30e7,  "D",       4.450647,  9,
  "Ca I",   1,      646.257,     4.70e7,  "D",       4.4409544, 7,
  "Ca I",   1,      649.378,     4.40e7,  "D",       4.4300117, 5,
  "Ca II",  2,      315.887,     3.10e8,  "C",       7.047169,  4,
  "Ca II",  2,      317.933,     3.60e8,  "C",       7.049551,  6,
  "Ca II",  2,      370.603,     8.80e7,  "C",       6.467875,  2,
  "Ca II",  2,      373.690,     1.70e8,  "C",       6.467875,  2
)
ionization_energy <- 6.1131549   # Ca I, eV
```

### Line intensities

`line_intensities()` finds each line within 0.2 nm of its tabulated
wavelength (`search`), which absorbs the offset of the wavelength axis, and
integrates it over ±0.15 nm (`half_width`). It keeps the columns of the
table of lines, so the result goes directly into a Boltzmann plot:

```{r intensities}
lines <- line_intensities(mean_spectrum, atomic)
lines |> select(species, wavelength, peak_wavelength, intensity, snr)
```

### Boltzmann and Saha-Boltzmann plots

```{r boltzmann}
neutral <- boltzmann_plot(filter(lines, stage == 1))
neutral
```

The Ca I lines come from upper levels between 3.9 and 4.8 eV. Over such a
narrow range, the slope is sensitive to the errors of a few intensities or
transition probabilities. The Saha equation places the lines of the ion on
the same plot, using the electron density, and extends the range of
energies to about 13 eV:

```{r saha}
saha <- saha_boltzmann_plot(lines, ionization_energy = ionization_energy, electron_density = ne)
saha
```

```{r saha-plot, fig.height = 4}
plot_boltzmann(saha)
```

The Saha-Boltzmann temperature, `r round(saha$temperature, -2)` K, is far
better determined than the Boltzmann temperature of the neutral lines
alone (`r round(neutral$temperature, -2)` ±
`r round(neutral$temperature_se, -1)` K). Its standard error,
`r round(saha$temperature_se, -1)` K, reflects the scatter of the points,
not the systematic errors of the atomic data or of the spectral response.

### Is the plasma in LTE?

The McWhirter criterion gives the minimum electron density for collisions
to dominate the population of the levels. The largest energy gap here is
the first excited level of Ca II, 3.15 eV above the ground state:

```{r mcwhirter}
lte <- mcwhirter_criterion(temperature = saha$temperature, delta_e = 3.15, electron_density = ne)
lte
```

The measured density is `r round(ne / lte$minimum_density)` times the
minimum. The criterion is necessary but not sufficient: in a transient,
inhomogeneous plasma, LTE also requires the equilibration to be faster
than the changes of the plasma (Cristoforetti et al., 2010).

## Self-absorption

Resonance lines end on the ground state, which is heavily populated, so
the plasma reabsorbs part of their light. `correct_self_absorption()`
compares each line with a reference line of the same species that is
little absorbed, by default the one of smallest optical depth (Sun and Yu,
2009). The potassium lines:

```{r self-absorption}
k_lines <- tribble(
  ~species, ~wavelength, ~Aki,      ~Ei,      ~Ek,      ~gk,
  "K I",    404.414,     1.150e+06, 0,        3.064907, 4,
  "K I",    404.721,     1.070e+06, 0,        3.062581, 2,
  "K I",    691.108,     2.500e+06, 1.609958, 3.403454, 2,
  "K I",    693.877,     4.956e+06, 1.617113, 3.403454, 2,
  "K I",    766.490,     3.779e+07, 0,        1.617113, 4,
  "K I",    769.896,     3.734e+07, 0,        1.609958, 2
)
line_intensities(mean_spectrum, k_lines) |>
  correct_self_absorption(temperature = saha$temperature) |>
  select(wavelength, measured_intensity, SA, reference)
```

The self-absorption coefficient `SA` is the fraction of the intensity that
escapes the plasma. The resonance doublet at 766.49 and 769.90 nm keeps only
a small fraction: the 766.49 nm line has twice the \(g_k A_{ki}\) of the
769.90 nm line, so it should be twice as intense, yet it is weaker. These
lines curve the calibration of `vignette("calibration")` and are unsuitable
for the calibration-free analysis below. The coefficients of the weaker
lines, between about 0.6 and 1.07 (hence the warning), show the precision
of the method: it inherits the noise of the reference line and the errors
of the transition probabilities.

## Calibration-free composition

Calibration-free LIBS (CF-LIBS; Ciucci et al., 1999) derives the
composition of a sample from its line intensities alone. In LTE, the
Boltzmann plots of all species share the same slope, \(-1/k_B T\), and the
intercept of each plot gives the density of its species. `cf_libs()` fits
the plots together, adds the ionization stages that are not observed with
the Saha equation, and normalizes the densities of the elements to
fractions that sum to one.

The lines must be unsaturated and optically thin, which rules out the
resonance lines. `cf_libs()` also needs the partition functions of the
species, which it computes from the energy levels of `nist_levels()`, and
the ionization energies:

```{r cf-fetch, eval = FALSE}
species <- c("Ca I", "Ca II", "Mg I", "Mg II", "K I", "K II", "Na I", "Na II")
levels <- nist_levels(species)
ionization <- nist_ionization_energy(c("Ca I", "Mg I", "K I", "Na I"))
```

To run offline, the atomic data and the partition functions (from
`partition_function()`, tabulated from 6000 to 12000 K) are included here:

```{r cf-atomic}
cf_atomic <- tribble(
  ~species, ~wavelength, ~Aki,      ~accuracy, ~Ek,      ~gk,
  "Ca I",   428.301,     4.340e+07, "C+",      4.779784, 5,
  "Ca I",   430.253,     1.360e+08, "C+",      4.779784, 5,
  "Ca I",   431.865,     7.400e+07, "C+",      4.769028, 3,
  "Ca I",   443.496,     6.700e+07, "C",       4.680635, 5,
  "Ca I",   445.478,     8.700e+07, "C",       4.681327, 7,
  "Ca I",   616.217,     4.770e+07, "C",       3.910399, 3,
  "Ca II",  315.887,     3.100e+08, "C",       7.047169, 4,
  "Ca II",  317.933,     3.600e+08, "C",       7.049551, 6,
  "Ca II",  370.603,     8.800e+07, "C",       6.467875, 2,
  "Ca II",  373.690,     1.700e+08, "C",       6.467875, 2,
  "Mg I",   383.829,     1.610e+08, "B+",      5.945916, 7,
  "Mg I",   516.732,     1.130e+07, "B+",      5.107827, 3,
  "Mg I",   517.268,     3.370e+07, "B+",      5.107827, 3,
  "Mg I",   518.360,     5.610e+07, "A",       5.107827, 3,
  "Mg II",  279.078,     4.010e+08, "A",       8.863762, 4,
  "Mg II",  279.800,     4.790e+08, "A",       8.863654, 6,
  "Mg II",  292.863,     1.150e+08, "A",       8.654711, 2,
  "Mg II",  293.651,     2.300e+08, "A",       8.654711, 2,
  "K I",    404.414,     1.150e+06, "A",       3.064907, 4,
  "K I",    404.721,     1.070e+06, "B+",      3.062581, 2,
  "K I",    691.108,     2.500e+06, "AA",      3.403454, 2,
  "K I",    693.877,     4.956e+06, "AA",      3.403454, 2,
  "Na I",   818.326,     4.290e+07, "A+",      3.616977, 4,
  "Na I",   819.482,     5.140e+07, "A+",      3.616971, 6
)
cf_ionization <- c(Ca = 6.113155, Mg = 7.646236, K = 4.340664, Na = 5.139077)

partition_table <- tribble(
  ~species, ~`6000`, ~`7000`, ~`8000`, ~`9000`, ~`10000`, ~`11000`, ~`12000`,
  "Ca I",   1.407,   1.798,   2.401,   3.281,   4.499,    6.184,    8.356,
  "Ca II",  2.389,   2.633,   2.917,   3.229,   3.560,    3.905,    4.262,
  "Mg I",   1.049,   1.109,   1.207,   1.362,   1.597,    1.959,    2.463,
  "Mg II",  2.001,   2.004,   2.010,   2.020,   2.036,    2.058,    2.087,
  "K I",    2.488,   3.000,   3.858,   5.078,   6.743,    8.938,    11.503,
  "K II",   1,       1,       1,       1,       1,        1,        1,
  "Na I",   2.166,   2.401,   2.868,   3.635,   4.779,    6.548,    8.724,
  "Na II",  1,       1,       1,       1,       1,        1,        1
)
# log-linear interpolation of the table, in the form cf_libs() accepts
partition <- function(species, temperature) {
  grid <- as.numeric(names(partition_table)[-1])
  vapply(species, function(s) {
    u <- unlist(partition_table[partition_table$species == s, -1])
    exp(stats::approx(grid, log(u), xout = temperature)$y)
  }, numeric(1))
}
```

With `partition = NULL` (the default), `cf_libs()` downloads the levels
and computes the partition functions itself. On the mean spectrum:

```{r cf-mean}
cf_mean <- line_intensities(mean_spectrum, cf_atomic) |>
  cf_libs(electron_density = ne, partition = partition, ionization_energy = cf_ionization)
cf_mean
```

```{r cf-plot, fig.height = 4.5}
plot_boltzmann(cf_mean)
```

The four Saha-Boltzmann plots are parallel within the scatter of their
points. The forage is mostly C, H, O and N, whose lines are not used here,
so the fractions describe the relative composition of Ca, Mg, K and Na,
which we compare with the laboratory values on the same basis:

```{r cf-compare}
laboratory <- forageLIBS |>
  summarise(across(c(Ca, Mg, K, Na), \(v) mean(v, na.rm = TRUE))) |>
  unlist()
cf_mean$composition |>
  select(element, cf_libs = mass_fraction) |>
  mutate(laboratory = laboratory[element] / sum(laboratory),
         ratio = cf_libs / laboratory)
```

```{r cf-numbers, include = FALSE}
ratios <- cf_mean$composition |>
  mutate(laboratory = laboratory[element] / sum(laboratory), ratio = mass_fraction / laboratory)
ratio_of <- \(e) round(ratios$ratio[ratios$element == e], 1)
```

Without any calibration, the composition is of the right order: the ratios
to the laboratory values are `r ratio_of("Ca")` for Ca,
`r ratio_of("Mg")` for Mg, `r ratio_of("K")` for K and `r ratio_of("Na")`
for Na. The main sources of bias are the spectral response of the
instrument, which weights the lines of each element differently, the
uncertain transition probabilities of the Ca lines, and the weak Na lines.
With one standard of known composition, the remaining bias can be measured
and removed for each element.

## Summary

-   **Screen for saturation first**: saturated lines are useless for
    diagnostics and nonlinear for calibration.
-   **Measure \(N_e\) on H\(\alpha\)**, a broad and optically thin line.
-   **Prefer Saha-Boltzmann to Boltzmann plots**: two ionization stages span
    a much wider range of energies.
-   **Check LTE and self-absorption** before trusting intensities: the
    McWhirter criterion is a minimum requirement, and resonance lines are
    often strongly self-absorbed.
-   **Calibration-free LIBS is semi-quantitative** without a correction of
    the spectral response.

## References

-   Ciucci, A., Corsi, M., Palleschi, V., Rastelli, S., Salvetti, A.,
    Tognoni, E. (1999). New procedure for quantitative elemental analysis
    by laser-induced plasma spectroscopy. *Applied Spectroscopy*,
    53(8):960–964.
-   Cristoforetti, G., De Giacomo, A., Dell'Aglio, M., Legnaioli, S.,
    Tognoni, E., Palleschi, V., Omenetto, N. (2010). Local thermodynamic
    equilibrium in laser-induced breakdown spectroscopy: beyond the
    McWhirter criterion. *Spectrochimica Acta Part B*, 65(1):86–95.
-   Gigosos, M.A., González, M.Á., Cardeñoso, V. (2003). Computer simulated
    Balmer-alpha, -beta and -gamma Stark line profiles for non-equilibrium
    plasmas diagnostics. *Spectrochimica Acta Part B*, 58(8):1489–1504.
-   Kramida, A., Ralchenko, Yu., Reader, J., and NIST ASD Team. NIST Atomic
    Spectra Database, <https://physics.nist.gov/asd>.
-   Sun, L., Yu, H. (2009). Correction of self-absorption effect in
    calibration-free laser-induced breakdown spectroscopy by an internal
    reference method. *Talanta*, 79(2):388–395.
