---
title: "Polynomial Baseline correction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polynomial Baseline correction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(include = FALSE)
source("~/Documents/Packages/specProc/R/baselinerm.R")
source("~/Documents/Packages/specProc/R/plotSpec.R")
```

```{r}
ssh = suppressPackageStartupMessages
ssh(library(tidyverse))
ssh(library(magrittr))
library(patchwork)
```

## LIBS spectrum in the 375–510 nm wavelength range

The plot below shows a typical LIBS spectrum. Prominent atomic and ionic
emission lines of Ca, Mn and Ba were identified using the [NIST spectral lines database](https://www.nist.gov/pml/atomic-spectra-database). The
spectrum show emission lines of Ca II 393.37 nm, Ca II 396.85 nm and Ca
I 422.67 nm, unresolved Mn triplet at 403.08, 403.31 and 403.45 nm, and
Ba ionic lines at 455.40 and 493.41 nm.

```{r}
data("Ca_Mn_spec", package = "specProc")
```

```{r}
spec_tbl <- Ca_Mn_spec %>% as_tibble()
```

```{r fig.height=4, fig.width=6}
spec_tbl %>% plotSpec()
```

Note that we used the ggplot-based `plotSpec` function to generate the
above plot.

## Baseline removal

When analyzing LIBS spectra, it is often more effective to subtract an
estimated baseline from the data. The estimate is constructed by fitting
a low-order polynomial function to the spectrum baseline. Then the
resulting curve fit is subtracted from the data. Here we use the
function `baselinerm` to perform such a task.

```{r message=FALSE, warning=FALSE}
baseline_fit <- spec_tbl %>%
  select(`390.03027`:`500.03397`) %>%
  baselinerm(degree = 7)
```

As one can see, the `baselinerm` function provides a list containing two
data frames, one for the baseline-subtracted spectrum or spectra `spec`
and the other for the fitted baseline `bkg`.

```{r}
str(baseline_fit, list.len = 5) 
```

We can then extract each data frame from the list using the
`purrr::pluck` function.

```{r}
background <- baseline_fit %>% pluck("bkg")
corrected_spec <- baseline_fit %>% pluck("spec")
```

And then visualize the spectrum before and after baseline removal.

```{r}
plot1 <- spec_tbl %>% 
  plotSpec() + 
  geom_line(
    data = background %>% 
      pivot_longer(cols = everything(), names_to = "wavelength", values_to = "intensity") %>%
      modify_at("wavelength", as.numeric), 
    aes(x = wavelength, y = intensity), 
    colour = "red"
  )
```

```{r}
plot2 <- corrected_spec %>% 
  plotSpec() + 
  geom_hline(yintercept = 0, colour = "red")
```

```{r fig.height=5, message=FALSE, warning=FALSE}
plot1 / plot2
```
