---
title: "Preprocessing LIBS spectra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing LIBS spectra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  dpi = 96
)
```

This vignette takes the raw spectra of the `specLIBS` data set through a
preprocessing pipeline: baseline correction, normalization, screening
for outlying shots, and averaging of replicates. Every step is a
modeling choice. For each one, the vignette shows how to check the choice
against the data rather than assuming it.

```{r setup}
library(specProc)
data(specLIBS)

meta <- specLIBS[1:8]
X <- as.matrix(specLIBS[-(1:8)])
wl <- as.numeric(colnames(X))
```

## The data and its structure

`specLIBS` contains `r nrow(X)` spectra of `r length(unique(meta$Sample))`
soil samples. Each sample was measured at 8 locations, and each spectrum
has `r ncol(X)` channels between `r round(min(wl))` and `r round(max(wl))`
nm, stored as raw detector counts.

```{r design}
table(locations_per_sample = table(meta$Sample))
summary(as.vector(X))
```

Two features of the data shape everything that follows:

-   **Replicates are nested within samples.** The 8 spectra of a sample share
    its composition and differ only by shot-to-shot variability and
    small-scale heterogeneity. Statistics that treat them as independent
    overstate precision.
-   **The counts include a large detector offset.** The smallest count is
    about 630, not zero. Any ratio, relative standard deviation or
    normalization computed on uncorrected counts is diluted by this offset.

We track five emission lines throughout, measured as the summed
intensity within ±0.15 nm of the line center:

```{r lines}
lines <- c(`Mg II 279.55` = 279.55, `Si I 288.16` = 288.16, `Ca II 393.37` = 393.37,
           `Al I 396.15` = 396.15, `K I 766.49` = 766.49)

line_areas <- function(M) {
  sapply(lines, function(center) rowSums(M[, abs(wl - center) < 0.15, drop = FALSE]))
}
```

## Step 1: Baseline correction

`baseline_arpls()` estimates the continuum with asymmetrically reweighted
penalized least squares. Its smoothing parameter `lambda` sets how stiff
the baseline is: small values follow narrow features and can eat into
emission lines, while large values miss curvature of the continuum.

```{r baseline-compare, fig.height = 4.5}
zoom <- wl > 385 & wl < 400
spec <- X[1, ]
fits <- sapply(c(1e2, 1e4, 1e6), function(l) {
  unlist(baseline_arpls(matrix(spec, 1), lambda = l, max.iter = 20)$background)
})

matplot(wl[zoom], cbind(spec[zoom], fits[zoom, ]), type = "l", lty = c(1, 1, 2, 3),
        col = c("grey40", "red", "blue", "darkgreen"), lwd = c(1, 2, 2, 2),
        ylim = c(700, 3000), xlab = "Wavelength [nm]", ylab = "Counts (lines truncated)")
legend("topleft", c("spectrum", "lambda = 1e2", "lambda = 1e4", "lambda = 1e6"),
       col = c("grey40", "red", "blue", "darkgreen"), lty = c(1, 1, 2, 3), lwd = 2, bty = "n")
```

With `lambda = 1e2`, the baseline rises into the base of the Ca and Al
lines and removes part of their area. Since the baseline is an estimate,
check how sensitive the quantity you care about is to `lambda`:

```{r lambda-sensitivity}
sens <- sapply(c(1e3, 1e4, 1e5, 1e6, 1e7), function(l) {
  bg <- as.matrix(baseline_arpls(X[1:8, ], lambda = l, max.iter = 20)$background)
  colMeans(line_areas(X[1:8, ] - bg))
})
colnames(sens) <- paste0("lambda=", c("1e3", "1e4", "1e5", "1e6", "1e7"))
round(sens)
```

Across four orders of magnitude of `lambda`, the areas of the strong
lines change by less than 2%. The weak K line, which sits on a relatively
larger continuum, changes by about 4%. The line areas are therefore robust
to this choice for strong lines, but not entirely for weak ones. We use
`lambda = 1e5`, the smallest value at which the baseline stays clear of the
line wings in the plot above. For your own data, run the same check on a
few representative spectra.

```{r baseline}
Xb <- as.matrix(baseline_arpls(X, lambda = 1e5, max.iter = 20)$correction)
```

The alternatives are `baseline_als()` (asymmetric least squares, with an
explicit asymmetry parameter `p`) and `baseline_lsp()` (iterative
polynomial fitting). All three return the corrected spectra together
with the estimated background.

## Step 2: Normalization

Normalization aims to remove multiplicative, shot-to-shot fluctuations in
the amount of ablated material and plasma conditions. It's often judged
by the relative standard deviation (RSD) of lines across replicates alone.
That criterion is incomplete, because a transformation that shrinks every
spectrum toward a common shape also reduces RSD, while destroying the
between-sample differences we want to measure.

A better criterion compares both sources of variation. The **intraclass
correlation** (ICC) is the fraction of total variance that lies between
samples rather than between replicates of a sample. The higher it is,
the better a line discriminates between samples relative to its
measurement noise. We estimate it from a one-way random-effects ANOVA:

```{r icc}
icc <- function(v, group) {
  if (stats::var(v) == 0) return(NA_real_)
  fit <- stats::anova(stats::lm(v ~ factor(group)))
  k <- mean(table(group))
  msb <- fit[1, "Mean Sq"]
  msw <- fit[2, "Mean Sq"]
  (msb - msw) / (msb + (k - 1) * msw)
}
rsd <- function(v, group) {
  median(tapply(v, group, function(z) sd(z) / mean(z) * 100))
}
```

We compare no normalization, total-area normalization, SNV, MSC, and
internal standardization to the Si I 288.16 nm line:

```{r normalizations}
si_cols <- colnames(X)[abs(wl - 288.16) < 0.15]
candidates <- list(
  `baseline only` = Xb,
  `area` = as.matrix(normalize(as.data.frame(Xb), method = "area")),
  `SNV` = as.matrix(snv(Xb)$correction),
  `MSC` = as.matrix(msc(Xb)$correction),
  `internal std. (Si)` = as.matrix(normalize(as.data.frame(Xb), method = "internal",
                                             wlength = si_cols))
)

evaluate <- function(M) {
  A <- line_areas(M)
  data.frame(
    line = colnames(A),
    RSD = apply(A, 2, rsd, group = meta$Sample),
    ICC = apply(A, 2, icc, group = meta$Sample)
  )
}
scores <- do.call(rbind, lapply(names(candidates), function(nm) {
  cbind(normalization = nm, evaluate(candidates[[nm]]))
}))
```

```{r normalization-table}
wide <- function(stat) {
  out <- tapply(scores[[stat]], list(scores$normalization, scores$line), identity)
  round(out[names(candidates), names(lines)], 2)
}
wide("RSD")   # median within-sample RSD (%)
wide("ICC")   # fraction of variance between samples
```

The two criteria disagree, and the disagreement is informative:

-   **SNV and MSC give the best repeatability.** They reduce the
    replicate RSD of the Mg, Ca and Al lines by about 40% compared with baseline
    correction alone.
-   **They also lower the ICC.** A lower ICC means the between-sample
    variance shrank even more than the within-sample variance. Normalization
    removes multiplicative variation, and part of the multiplicative
    variation between samples is systematic: soils of different texture
    ablate and emit differently.
-   **Replicates alone cannot say whether that variation is useful.** If the
    between-sample intensity differences are a matrix effect unrelated to the
    property of interest, removing them helps. If they carry information
    about it, removing them hurts. Only a validated model of that property
    can decide; the calibration vignette does this for clay content.
-   **The weak K line is a special case.** SNV increases its RSD: dividing
    by the whole-spectrum standard deviation adds that statistic's noise to
    a line that is itself noisy.
-   **Internal standardization by Si is not usable here.** It sets the Si
    line to a constant, so its own RSD and ICC are meaningless. It also
    assumes the silicon content is constant across samples, which is false
    for soils ranging from clay to sand. It lowers the ICC of every other
    line.

No normalization is universally best, so choose one on data from the
matrix at hand, and confirm the choice against the end goal. We continue
with SNV, the most repeatable option, and revisit the choice in the
calibration vignette:

```{r snv}
Xn <- candidates$SNV
```

## Step 3: Screening outlying shots

A misfired or defocused shot can bias a sample's mean spectrum. Outliers
must be judged relative to the other shots of the *same sample*, because
the differences between samples are real signal. We therefore remove
each sample's median from its line intensities, and look for shots that
are outlying in this within-sample deviation space.

```{r deviations}
A <- line_areas(Xn)
deviation <- A - apply(A, 2, function(v) ave(v, meta$Sample, FUN = median))
colnames(deviation) <- sub(" .*", "", colnames(deviation))
```

`plot_outliers()` computes robust (MCD) Mahalanobis distances, which
account for the correlation between lines. It then displays the
standardized deviation of every shot for each variable:

```{r outlier-plot, fig.height = 4.5}
plot_outliers(deviation, quan = 0.75, show.mahal = TRUE)
```

The default cutoff of `plot_outliers()` flags a large share of the shots:

```{r outlier-table}
screen <- plot_outliers(deviation, quan = 0.75, show.outlier = FALSE, show.mahal = FALSE)
table(flagged = screen$outlier)
```

That is about 11% of the shots, far more than the few gross errors one
would expect from misfires. Before discarding anything, compare the
distribution of the squared robust distances with the χ² distribution
they would follow if the within-sample deviations were multivariate normal:

```{r qq, fig.height = 4.5, fig.width = 5, out.width = "65%"}
d2 <- screen$mahalanobis^2
p <- ncol(deviation)
qqplot(stats::qchisq(stats::ppoints(length(d2)), df = p), d2, log = "xy",
       xlab = expression(chi[5]^2 ~ "quantiles (log scale)"),
       ylab = "Squared robust distance (log scale)",
       main = "Robust distances vs the normal model")
abline(0, 1, col = "red")
abline(h = stats::qchisq(0.999, p), lty = 2)
```

The bulk of the shots follows the reference line: the median distance
matches its χ² expectation. The upper tail is much heavier than the normal
model predicts. Most of these shots are not errors: they reflect genuine
heterogeneity of the soil at the scale of the laser spot. Discarding them
would bias each sample's mean toward its most homogeneous spots.

A defensible rule is to remove only the gross outliers, beyond the 99.9%
quantile of the χ² distribution, and to report how many shots were removed:

```{r screen}
flagged <- d2 > stats::qchisq(0.999, p)
sum(flagged)
table(shots_kept = table(meta$Sample[!flagged]))
sort(table(meta$Sample[flagged]), decreasing = TRUE)[1:5]
```

Most samples keep all 8 shots. The few samples that lose several shots
are heterogeneous and deserve inspection: their mean spectrum rests on
fewer measurements and is less certain.

## Step 4: Averaging replicates

With the screened shots, `average()` reduces each sample to one mean
spectrum. It is implemented in C++ and handles the full matrix in a
fraction of a second:

```{r average}
Xs <- average(cbind(Sample = meta$Sample[!flagged], as.data.frame(Xn[!flagged, ])), Sample)
dim(Xs)
```

The mean of the remaining shots still includes the heavy-tailed but
legitimate variability discussed above. If you prefer not to screen at
all, the per-sample median of each channel is a robust alternative to the
mean, but it is less efficient when the data are close to normal.

## The result

The preprocessed, sample-level data set is ready for exploratory analysis
or calibration:

```{r final, fig.height = 4}
samples <- meta[match(Xs$Sample, meta$Sample), ]
by_type <- average(cbind(Type = samples$Type, Xs[-1]), Type)
keep <- names(by_type)[-1][wl > 380 & wl < 400]
plot_spectra(by_type[c("Type", keep)], id = Type) +
  ggplot2::theme(legend.position = "top") +
  ggplot2::labs(color = NULL, y = "SNV intensity", title = "Mean preprocessed spectrum by soil type")
```

The companion vignettes use this pipeline to fit emission lines
(`vignette("line-fitting", package = "specProc")`) and to build a
calibration model for clay content
(`vignette("calibration", package = "specProc")`).
