---
title: "2. Preprocessing with recipe steps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Preprocessing with recipe steps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%",
  dpi = 96
)
```

```{r requirements, include = FALSE}
needed <- c("recipes", "parsnip", "plsmod", "mixOmics", "workflows", "tune", "rsample", "yardstick")
if (!all(vapply(needed, requireNamespace, logical(1), quietly = TRUE))) {
  knitr::knit_exit("This vignette needs the tidymodels packages recipes, parsnip, plsmod (with mixOmics), workflows, tune, rsample and yardstick.")
}
```

Preprocessing turns raw spectra into variables ready for modeling. In
specProc, each preprocessing method is also a
[recipes](https://recipes.tidymodels.org) step (`step_*()`). A recipe is
estimated on training data (`prep()`) and applied unchanged to new
spectra (`bake()`), and within a tidymodels workflow it is re-estimated in
every resample and can be tuned with the model. This vignette builds a
recipe for the forage spectra, and uses it to predict potassium.

```{r setup, message = FALSE}
suppressPackageStartupMessages(library(mixOmics))
library(specProc)
library(recipes)
library(parsnip)
library(plsmod)
library(workflows)
library(tune)
library(rsample)
library(yardstick)
library(dplyr)
library(purrr)
library(ggplot2)
# mixOmics (the PLS engine) and MASS, which it attaches, have functions
# named like those of the tidyverse and tidymodels: use the latter
select <- dplyr::select
map <- purrr::map
tune <- tune::tune

data("forageLIBS")
spectra_id <- forageLIBS |> select(1:2) |> names()
minerals <- forageLIBS |> select(3:14) |> names()
spectra <- forageLIBS |> select(-all_of(c(spectra_id, minerals)))
wl <- as.numeric(names(spectra))
```

The recipe predicts K from the spectral channels. The identifiers and the
other elements keep a role of their own, so they are carried along but not
used as predictors:

```{r recipe}
base <- recipe(K ~ ., data = forageLIBS) |>
  update_role(all_of(spectra_id), all_of(setdiff(minerals, "K")), new_role = "id")
```

## Baseline correction

The continuum emission of the plasma and the detector offset add a smooth
background to every spectrum. `step_baseline()` removes it, by default with
asymmetrically reweighted penalized least squares (`baseline_arpls()`).
Its parameter `lambda` sets the stiffness of the baseline: too small, it
cuts into the lines; too large, it misses the curvature of the continuum.
`baseline_arpls()` shows the baseline itself:

```{r baseline, fig.height = 3.5}
region <- wl > 276 & wl < 290
fitted_baselines <- map(c(`1e3` = 1e3, `1e5` = 1e5, `1e7` = 1e7), \(lambda) {
  unlist(baseline_arpls(spectra[1, ], lambda = lambda, max.iter = 20)$background)[region]
})
first_spectrum <- tibble(wavelength = wl[region], counts = unlist(spectra[1, region]))

tibble(wavelength = wl[region], !!!fitted_baselines) |>
  tidyr::pivot_longer(-wavelength, names_to = "lambda", values_to = "counts") |>
  ggplot(aes(wavelength, counts)) +
  geom_line(data = first_spectrum, colour = "grey55") +
  geom_line(aes(colour = lambda), linewidth = 0.8) +
  coord_cartesian(ylim = c(800, 2500)) +
  labs(x = "Wavelength (nm)", y = "Counts (lines truncated)") +
  theme_bw() +
  theme(legend.position = "top")
```

With `lambda = 1e7`, the baseline is too stiff to follow the broad hump of
the continuum under the Mg lines near 280 nm; with `1e3`, it starts to rise
into the lines. `1e5` follows the continuum without touching them.
`lambda` can also be tuned with the model (`lambda = tune()`, with the
range of `baseline_lambda()`).

```{r step-baseline}
corrected <- base |>
  step_baseline(all_predictors(), lambda = 1e5, options = list(max.iter = 20))
```

## Normalization

The intensity of every line changes from spectrum to spectrum with the
ablated mass and the plasma conditions. Normalization divides each
spectrum by a quantity that follows these changes:

| Step | Divides each spectrum by |
|---|---|
| `step_spectral_norm()` | its total area, or its L1, L2 or maximum norm |
| `step_snv()` | its standard deviation, after centering (standard normal variate) |
| `step_msc()` | a regression on the mean spectrum (multiplicative scatter correction) |
| `step_emsc()` | an extended model with a polynomial baseline |
| `step_line_ratio()` | the intensity of a reference line (internal standard) |

A good normalization brings the line intensities closer to the
concentrations. `step_line_intensities()` replaces the channels by the
intensities of chosen lines, so we compare the normalizations by the
correlation of each line with the laboratory concentration of its element:

```{r normalizations}
lines <- c(K = 769.896, Mg = 285.213, Ca = 317.933, Na = 589.592, Mn = 257.610, Fe = 259.940)
normalizations <- list(
  none = identity,
  area = \(r) step_spectral_norm(r, all_predictors(), method = "area"),
  SNV = \(r) step_snv(r, all_predictors()),
  MSC = \(r) step_msc(r, all_predictors()),
  `C I 247.86` = \(r) step_line_ratio(r, all_predictors(), reference = 247.856, window = 0.15)
)

imap(normalizations, \(normalize, name) {
  intensities <- corrected |>
    normalize() |>
    step_line_intensities(all_predictors(), lines = lines) |>
    prep() |>
    bake(new_data = NULL)
  map_dbl(names(lines), \(e) cor(intensities[[paste0("line_", e)]], forageLIBS[[e]],
                                 use = "complete.obs")) |>
    set_names(names(lines))
}) |>
  bind_rows(.id = "normalization") |>
  mutate(across(-normalization, \(r) round(r, 2)))
```

Normalizing to the carbon line C I 247.86 nm gives the highest
correlations for K, Mg and Ca. Carbon is the main element of the organic
matrix, and its content varies little between forages: it is an internal
standard. The other normalizations help less. The correlations of Na, Mn
and Fe change little: their concentrations vary over a much wider range
than the normalization factor.

## Other steps

| Purpose | Steps |
|---|---|
| Smoothing, derivatives | `step_savgol()` |
| Compression | `step_wavelet()` (wavelet coefficients), `step_line_intensities()` (line intensities) |
| Scaling | `step_pareto_scale()`, `step_poisson_scale()` |
| Removing unwanted variation | `step_epo()`, `step_glsw()`, `step_osc()`, `step_direct_osc()`, `step_projected_osc()`, `step_opls()`, `step_o2pls()`, `step_y_gradient_glsw()` |
| Robust transformations | `step_robust_bcyj()`, `step_robpca()`, `step_rospca()`, `step_macropca()`, `step_cellpca()` |
| Replicate shots | `step_reject_shots()` |

For example, `step_wavelet()` keeps the approximation coefficients of a
discrete wavelet transform, a smoothed spectrum with 8 times fewer values
at `level = 3`:

```{r wavelet}
corrected |>
  step_wavelet(all_predictors(), level = 3) |>
  prep() |>
  bake(new_data = NULL) |>
  select(starts_with("wav_")) |>
  ncol()
```

Steps that use the outcome, such as `step_opls()` or `step_osc()`, must be
estimated on training data only. Within a workflow, this is automatic.

## A model for potassium

We keep 20% of the samples for testing. The data hold a few repeated
samples, so the splits keep the spectra of a sample together:

```{r split}
set.seed(1)
split <- group_initial_split(forageLIBS, group = Sample, prop = 0.8)
training_set <- training(split)
folds <- group_vfold_cv(training_set, group = Sample, v = 5)
```

The model is a partial least squares (PLS) regression, from the mixOmics
engine of parsnip, with its number of components tuned. We compare three
recipes: baseline correction followed by SNV, by the carbon internal
standard, or by the carbon standard and a wavelet compression:

```{r recipes}
baseline_step <- \(r) step_baseline(r, all_predictors(), lambda = 1e5, options = list(max.iter = 20))
carbon_step <- \(r) step_line_ratio(r, all_predictors(), reference = 247.856, window = 0.15)

recipes <- list(
  SNV = base |> baseline_step() |> step_snv(all_predictors()),
  carbon = base |> baseline_step() |> carbon_step(),
  `carbon + wavelet` = base |> baseline_step() |> carbon_step() |>
    step_wavelet(all_predictors(), level = 3)
)

# mixOmics scales every predictor by default; scale = FALSE only centers
pls_model <- pls(num_comp = tune()) |>
  set_mode("regression") |>
  set_engine("mixOmics", scale = FALSE)
```

```{r tune}
tuned <- map(recipes, \(rec) {
  tune_grid(workflow(rec, pls_model), resamples = folds,
            grid = tibble(num_comp = c(2, 4, 6, 8, 10, 12, 15)),
            metrics = metric_set(rmse, rsq))
})

map(tuned, \(res) show_best(res, metric = "rmse", n = 1)) |>
  bind_rows(.id = "recipe") |>
  select(recipe, num_comp, rmse = mean, std_err)
```

```{r tune-plot, fig.height = 3.5}
map(tuned, collect_metrics) |>
  bind_rows(.id = "recipe") |>
  filter(.metric == "rmse") |>
  ggplot(aes(num_comp, mean, colour = recipe)) +
  geom_line() +
  geom_point() +
  labs(x = "PLS components", y = "Cross-validated RMSE (% K)", colour = NULL) +
  theme_bw()
```

```{r best, include = FALSE}
scores <- map_dbl(tuned, \(res) show_best(res, metric = "rmse", n = 1)$mean)
best_recipe <- names(which.min(scores))
```

The three recipes give similar errors once the number of components is
tuned: PLS finds the potassium information in all of them, and the
normalization matters less for a multivariate model than for a single
line. The `r best_recipe` recipe is the best by a small margin. We refit
it on the whole training set with its best number of components, and
evaluate it once on the test samples:

```{r final}
final <- workflow(recipes[[best_recipe]], pls_model) |>
  finalize_workflow(select_best(tuned[[best_recipe]], metric = "rmse")) |>
  last_fit(split, metrics = metric_set(rmse, rsq))
collect_metrics(final)
```

```{r test-numbers, include = FALSE}
test_rmse <- collect_metrics(final)$.estimate[1]
cv_rmse <- min(scores)
```

The test error, `r round(test_rmse, 2)`% K, is somewhat larger than the
cross-validated one (`r round(cv_rmse, 2)`%): with
`r nrow(testing(split))` test spectra, it is itself uncertain.

```{r final-plot, fig.height = 4, fig.width = 4.5, out.width = "60%"}
collect_predictions(final) |>
  ggplot(aes(K, .pred)) +
  geom_abline(colour = "grey50", linetype = "dashed") +
  geom_point(alpha = 0.6) +
  coord_equal() +
  labs(x = "Laboratory K (%)", y = "Predicted K (%)") +
  theme_bw()
```

The fitted workflow applies every step, with the values estimated on the
training data, to new spectra: `predict(extract_workflow(final), new_data)`.
The prepped recipe also shows what each step estimated, with `tidy()`.
The next vignette, `vignette("calibration")`, builds univariate
calibration curves and computes the figures of merit of both kinds of
models.
